### 2024-12-01 other/datasheet/datasheet_template.qmd
Update datasheet_template.qmd

---
title: "Datasheet for 'A dataset'"
subtitle: "My subtitle if needed"
author: 
  - Karen Riani
thanks: "Code and data are available at: https://github.com/karenrni/Analyzing-Torontos-Speeders."
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable analysis of Australian politicians. We were unable to find a publicly available dataset in a structured format that had the biographical and political information on Australian politicians that was needed for modelling.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - TBD
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - TBD
4. *Any other comments?*
    - TBD

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- TBD
2. *How many instances are there in total (of each type, if appropriate)?*
	- TBD
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- TBD
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- TBD
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- TBD
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- TBD
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- TBD
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- TBD
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- TBD
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- TBD
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- TBD
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- TBD
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- TBD
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- TBD
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- TBD
16. *Any other comments?*
	- TBD

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- TBD
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- TBD
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- TBD
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- TBD
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- TBD
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- TBD
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- TBD
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- TBD
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- TBD
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
12. *Any other comments?*
	- TBD

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- TBD
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- TBD
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- TBD
4. *Any other comments?*
	- TBD

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- TBD
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- TBD
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- TBD
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- TBD
6. *Any other comments?*
	- TBD

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- TBD
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- TBD
3. *Is there an erratum? If so, please provide a link or other access point.*
	- TBD
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- TBD
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- TBD
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- TBD
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
8. *Any other comments?*
	- TBD

\newpage

# References

### 2024-12-01 paper/paper.qmd
Update paper.qmd

---
title: "Chasing Speed Demons: Analyzing Toronto’s Speeding Hotspots and Enforcement Gaps”"
subtitle: "My subtitle if needed"
author: 
  - Karen Riani
thanks: "Code and data are available at: https://github.com/karenrni/Analyzing-Torontos-Speeders/."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @talia, @opendatatorotno, and @rohan.

The remainder of this paper is structured as follows. @sec-mydatasection

# Data {#sec-mydatasection}

Some of our data is of penguins (@fig-marriage), from @palmerpenguins.

```{r}
#| label: fig-marriage
#| fig-cap: Bills of penguins
#| echo: false

data |> 
  ggplot(aes(x = date, y = marriage_licenses)) +
  geom_point()

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

\newpage

# References

### 2024-12-01 scripts/00-simulate_data.R
preamble updates

@@ -1,19 +1,17 @@
 #### Preamble ####
 # Purpose: Simulates data
-# Author: Rohan Alexander
-# Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
+# Author: Karen Riani
+# Date: 1 December 2024
+# Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: None
 # Any other information needed? None
-
 
 #### Workspace setup ####
 library(tidyverse)
 
 
 #### Simulate data ####
-set.seed(304)
+set.seed(777)
 
 # Define the start and end date
 start_date <- as.Date("2018-01-01")

### 2024-12-01 scripts/01-download_data.R

@@ -1,9 +1,8 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Open Data Toronto
-# Author: Rohan Alexander
-# Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
+# Author: Karen Riani
+# Date: 1 December 2024
+# Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: None
 # Any other information needed? None
 

### 2024-12-01 scripts/02-data_cleaning.R

@@ -1,9 +1,8 @@
 #### Preamble ####
 # Purpose: Cleans the raw marriage data into an analysis dataset
-# Author: Rohan Alexander
-# Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
+# Author: Karen Riani
+# Date: 1 December 2024
+# Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: Need to have downloaded the data
 # Any other information needed? None.
 

### 2024-12-01 scripts/03-test_analysis_data.R

@@ -1,9 +1,8 @@
 #### Preamble ####
 # Purpose: Sanity check of the data
-# Author: Rohan Alexander
-# Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
+# Author: Karen Riani
+# Date: 1 December 2024
+# Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: Need to have simulated data
 # Any other information needed? None.
 

### 2024-12-01 scripts/04-model.R

@@ -1,11 +1,10 @@
 #### Preamble ####
 # Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Author: Karen Riani
+# Date: 1 December 2024
+# Contact: karen.riani@mail.utoronto.ca
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####

### 2024-12-01 scripts/01-download_data.R
download

@@ -11,21 +11,21 @@
 library(opendatatoronto)
 library(tidyverse)
 
-
 #### Download data ####
+
 # get package
-package <- show_package("e28bc818-43d5-43f7-b5d9-bdfb4eda5feb")
+package <- show_package("058236d2-d26e-4622-9665-941b9e7a5229")
+package
 
 # get all resources for this package
-resources <- list_package_resources("e28bc818-43d5-43f7-b5d9-bdfb4eda5feb")
+resources <- list_package_resources("058236d2-d26e-4622-9665-941b9e7a5229")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
 datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
 
 # load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% 
-  get_resource()
-
+data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+data
 
 #### Save data ####
 write_csv(data, "data/raw_data/raw_data.csv") 

### 2024-12-01 scripts/01-download_data.R
data download error- temp fix

@@ -28,4 +28,4 @@ data <- filter(datastore_resources, row_number()==1) %>% get_resource()
 data
 
 #### Save data ####
-write_csv(data, "data/raw_data/raw_data.csv") 
+write_csv(data, "data/raw_data/raw_data.csv")

### 2024-12-01 scripts/02-data_cleaning.R

@@ -10,7 +10,9 @@
 library(tidyverse)
 
 #### Clean data ####
-raw_data <- read_csv("data/raw_data/raw_data.csv")
+raw_data <- read_csv("data/raw_data/WYS_raw_data.csv")
+
+colnames(raw_data)
 
 cleaned_data <-
   raw_data |>

### 2024-12-02 scripts/01-download_data.R
update for reproduction

@@ -4,8 +4,7 @@
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: None
-# Any other information needed? None
-
+# Any other information needed: As of this date, the API download does not properly load the data and has issues with mismatched column lengths for topics and formats. Thus I provided a secondary option in case it does not work in the future. 
 
 #### Workspace setup ####
 library(opendatatoronto)
@@ -13,6 +12,10 @@ library(tidyverse)
 
 #### Download data ####
 
+### Option 1: Load through direct link to csv
+WYS_raw_data<- read_csv("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/058236d2-d26e-4622-9665-941b9e7a5229/resource/969583a4-4e9e-44e8-be79-a1b6f1dbd74e/download/WYS%20Mobile%20Sign%20Summary.csv")
+
+### Option 2: Load through API (Currently has issues with tibble)
 # get package
 package <- show_package("058236d2-d26e-4622-9665-941b9e7a5229")
 package

### 2024-12-02 scripts/01-download_data.R
update topic+ downloads

@@ -4,7 +4,7 @@
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: None
-# Any other information needed: As of this date, the API download does not properly load the data and has issues with mismatched column lengths for topics and formats. Thus I provided a secondary option in case it does not work in the future. 
+# Any other information needed: As of this date, the Speed Program API download does not properly load the data and has issues with mismatched column lengths for topics and formats. Thus I provided a secondary option in case it does not work in the future. 
 
 #### Workspace setup ####
 library(opendatatoronto)
@@ -12,23 +12,42 @@ library(tidyverse)
 
 #### Download data ####
 
-### Option 1: Load through direct link to csv
-WYS_raw_data<- read_csv("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/058236d2-d26e-4622-9665-941b9e7a5229/resource/969583a4-4e9e-44e8-be79-a1b6f1dbd74e/download/WYS%20Mobile%20Sign%20Summary.csv")
+### Speed program data ###
+## OPTION 1: Download csv directly from provided OpenDataToronto link
+speed_data_raw<- read_csv("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/642efeca-1258-4c23-8e86-d9affca26001/resource/866701b9-9e1f-4c39-92df-e1ec9e20cbbe/download/Stationary%20Sign%20locations%20-%204326.csv")
 
-### Option 2: Load through API (Currently has issues with tibble)
+
+##OPTION 2 (As of this date, this does not work): Download for Developers from OpenDataToronto
 # get package
-package <- show_package("058236d2-d26e-4622-9665-941b9e7a5229")
-package
+packageSpeed <- show_package("642efeca-1258-4c23-8e86-d9affca26001")
+packageSpeed
 
 # get all resources for this package
-resources <- list_package_resources("058236d2-d26e-4622-9665-941b9e7a5229")
+resourcesSpeed <- list_package_resources("642efeca-1258-4c23-8e86-d9affca26001")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
[…]

### 2024-12-02 scripts/02-data_cleaning.R
update cleaning-- only speed working.

@@ -8,20 +8,73 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(arrow)
+library(sf)
 
 #### Clean data ####
-raw_data <- read_csv("data/raw_data/WYS_raw_data.csv")
+# Read the raw data
+raw_data_cams <- read_csv("data/raw_data/raw_data_cameras.csv")
+raw_data_speeds <- read_csv("data/raw_data/raw_data_speeds.csv")
 
-colnames(raw_data)
+# Step 1: Remove rows with missing or invalid geometry
+raw_data_cams <- raw_data_cams %>%
+  filter(!is.na(geometry) & str_detect(geometry, "\\[\\[.*\\]\\]"))
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |> 
-  separate(col = time_period,
-            into = c("year", "month"),
-            sep = "-") |> 
-  mutate(date = lubridate::ymd(paste(year, month, "01", sep = "-"))
-         )
-  
-#### Save data ####
-write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")
+raw_data_speeds <- raw_data_speeds %>%
+  filter(!is.na(geometry) & str_detect(geometry, "\\[\\[.*\\]\\]"))
+
+# Step 2: Parse geometry column to extract coordinates
+extract_coordinates <- function(geometry) {
+  # Use regular expressions to extract the coordinate values from the JSON-like string
+  coords <- str_match(geometry, "\\[\\[(-?\\d+\\.\\d+),\\s*(-?\\d+\\.\\d+)\\]\\]")
+  tibble(longitude = as.numeric(coords[, 2]), latitude = as.numeric(coords[, 3]))
+}
+
+# Apply extraction to cameras data
+cams_coords <- raw_data_cams %>%
+  filter(!is.na(geometry)) %>%  # Remove rows with missing geometry
+  mutate(extracted = map(geometry, extract_coordinates)) %>%
+  unnest_wider(extracted) %>%
+  filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
+
+# Apply extraction to speeds data
+speeds_coords <- raw_data_speeds %>%
+  filter(!is.na(geometry)) %>%  # Remove rows with missing geometry
+  mutate(extracted = map(geometry, extract_coordinates)) %>%
+  unnest_wider(extracted) %>%
+  filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
+
[…]

### 2024-12-02 scripts/02-data_cleaning.R
update full cleaned data, fix for traffic cam data

@@ -12,30 +12,40 @@ library(arrow)
 library(sf)
 
 #### Clean data ####
+
 # Read the raw data
 raw_data_cams <- read_csv("data/raw_data/raw_data_cameras.csv")
 raw_data_speeds <- read_csv("data/raw_data/raw_data_speeds.csv")
 
-# Step 1: Remove rows with missing or invalid geometry
-raw_data_cams <- raw_data_cams %>%
-  filter(!is.na(geometry) & str_detect(geometry, "\\[\\[.*\\]\\]"))
+# Check the structure and sample values of the geometry column for cameras
+print(head(raw_data_cams$geometry))
+print(class(raw_data_cams$geometry))
 
-raw_data_speeds <- raw_data_speeds %>%
-  filter(!is.na(geometry) & str_detect(geometry, "\\[\\[.*\\]\\]"))
+#### Step 1: Handle camera data with geometry in c(...) format ####
+extract_camera_coordinates <- function(geometry) {
+  # Remove "c(" and ")" from the string, then split into longitude and latitude
+  coords <- str_remove_all(geometry, "c\\(|\\)") %>%
+    str_split(",\\s*") %>%
+    unlist() %>%
+    as.numeric()
+  tibble(longitude = coords[1], latitude = coords[2])
+}
 
-# Step 2: Parse geometry column to extract coordinates
+# Extract coordinates and create sf object for cameras
+cams_coords <- raw_data_cams %>%
+  filter(!is.na(geometry)) %>%  # Remove rows with missing geometry
+  mutate(extracted = map(geometry, extract_camera_coordinates)) %>%
+  unnest_wider(extracted) %>%  # Expand coordinates into longitude and latitude columns
+  filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
+
+cams_sf <- st_as_sf(cams_coords, coords = c("longitude", "latitude"), crs = 4326)
+
+#### Step 2: Parse geometry column to extract coordinates for speed data ####
 extract_coordinates <- function(geometry) {
   # Use regular expressions to extract the coordinate values from the JSON-like string
   coords <- str_match(geometry, "\\[\\[(-?\\d+\\.\\d+),\\s*(-?\\d+\\.\\d+)\\]\\]")
   tibble(longitude = as.numeric(coords[, 2]), latitude = as.numeric(coords[, 3]))
 }
-
-# Apply extraction to cameras data
[…]

### 2024-12-02 scripts/01-download_data.R
updates with speeder counts

@@ -1,37 +1,21 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from Open Data Toronto
+# Purpose: Downloads and processes data from Open Data Toronto
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: None
-# Any other information needed: As of this date, the Speed Program API download does not properly load the data and has issues with mismatched column lengths for topics and formats. Thus I provided a secondary option in case it does not work in the future. 
+# Any other information needed: As of this date, the Speed Program API download has issues with mismatched column lengths for topics and formats. A secondary option is provided for future compatibility.
 
 #### Workspace setup ####
 library(opendatatoronto)
 library(tidyverse)
+library(janitor)
 
 #### Download data ####
 
 ### Speed program data ###
-## OPTION 1: Download csv directly from provided OpenDataToronto link
-speed_data_raw<- read_csv("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/642efeca-1258-4c23-8e86-d9affca26001/resource/866701b9-9e1f-4c39-92df-e1ec9e20cbbe/download/Stationary%20Sign%20locations%20-%204326.csv")
-
-
-##OPTION 2 (As of this date, this does not work): Download for Developers from OpenDataToronto
-# get package
-packageSpeed <- show_package("642efeca-1258-4c23-8e86-d9affca26001")
-packageSpeed
-
-# get all resources for this package
-resourcesSpeed <- list_package_resources("642efeca-1258-4c23-8e86-d9affca26001")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resourcesSpeed <- filter(resourcesSpeed, tolower(format) %in% c('csv', 'geojson'))
-
-# load the first datastore resource as a sample
-speed_data_raw <- filter(datastore_resourcesSpeed, row_number()==1) %>% get_resource()
-speed_data_raw
-
+## Download csv directly from provided OpenDataToronto link
[…]

### 2024-12-02 scripts/02-data_cleaning.R
update for speeder count!!!

@@ -80,13 +80,20 @@ distance_threshold <- 500  # Distance in meters
 joined_data <- cams_sf %>%
   st_join(speeds_sf, join = st_is_within_distance, dist = distance_threshold)
 
-# View joined data
-print(head(joined_data))
-str(joined_data)
+# Track school zones without cameras
+joined_data <- joined_data %>%
+  mutate(no_camera_in_radius = ifelse(is.na(sign_id), TRUE, FALSE))  # Use `sign_id` to track camera presence
 
-#### Step 6: Save results ####
-# Drop the geometry column before saving, not supported by parquet
-joined_data_flat <- st_drop_geometry(joined_data)
+#### Step 6: Convert geometry back into longitude and latitude columns #### 
+joined_data <- joined_data %>%
+  mutate(longitude = st_coordinates(.)[, 1],  # Extract X (longitude)
+         latitude = st_coordinates(.)[, 2]) %>%  # Extract Y (latitude)
+  st_drop_geometry()  # Remove the geometry column
+
+#### Step 7: Remove unnecessary columns #### 
+colnames(joined_data)
+
+#### Step 8: Save results ####
 
 # Save the flattened data as a Parquet file
 arrow::write_parquet(joined_data_flat, "data/analysis_data/analysis_data.parquet")

### 2024-12-02 scripts/02-data_cleaning.R
finish cleaning, join speed count data

@@ -16,6 +16,7 @@ library(sf)
 # Read the raw data
 raw_data_cams <- read_csv("data/raw_data/raw_data_cameras.csv")
 raw_data_speeds <- read_csv("data/raw_data/raw_data_speeds.csv")
+speed_counts_data_raw <- read_csv("data/raw_data/raw_data_speeding_count.csv")
 
 # Check the structure and sample values of the geometry column for cameras
 print(head(raw_data_cams$geometry))
@@ -90,10 +91,28 @@ joined_data <- joined_data %>%
          latitude = st_coordinates(.)[, 2]) %>%  # Extract Y (latitude)
   st_drop_geometry()  # Remove the geometry column
 
-#### Step 7: Remove unnecessary columns #### 
-colnames(joined_data)
+#### Step 7: Use Stratified Random Sampling on Speed Count Data #### 
+speed_counts_data_cleaned <- speed_counts_data_raw %>%
+  select(sign_id, datetime_bin, speed_bin, volume) %>%  # Select relevant columns
+  group_by(sign_id) %>%          # Group by `sign_id`
+  sample_frac(0.01) %>%          # Take 1% per group
+  ungroup()
 
-#### Step 8: Save results ####
+#### Step 8: Join Speed Counts with Joined Camera and Speed data #### 
+# Perform an inner join to ensure no NA values in `sign_id`
+final_joined_data <- joined_data %>%
+  inner_join(speed_counts_data_cleaned, by = "sign_id", relationship = "many-to-many")  # Specify many-to-many relationship
 
-# Save the flattened data as a Parquet file
-arrow::write_parquet(joined_data_flat, "data/analysis_data/analysis_data.parquet")
+#### Step 9: Filter Dates and Select Specific Columns ####
+
+# Filter the data
+final_filtered_data <- final_joined_data %>%
+  filter(as.Date(end_date) >= as.Date("2023-01-01"),          # Keep rows with end_date in 2023 or later
+         !grepl("^2024", as.character(start_date))) %>%       # Exclude rows with start_date starting with 2024
+  select(sign_id, X_id, longitude, latitude,                 # Select relevant columns
+         speed_limit, volume, speed_bin, no_camera_in_radius)
+
+#### Step 10: Save Results ####
+
+# Save the final filtered data as a Parquet file
+arrow::write_parquet(final_filtered_data, "data/analysis_data/analysis_data.parquet")

### 2024-12-02 scripts/02-data_cleaning.R
logic error handling update + refs

@@ -81,6 +81,14 @@ distance_threshold <- 500  # Distance in meters
 joined_data <- cams_sf %>%
   st_join(speeds_sf, join = st_is_within_distance, dist = distance_threshold)
 
+# Ensure CRS is set for joined data
+if (is.na(st_crs(joined_data))) {
+  joined_data <- st_set_crs(joined_data, 32617)
+}
+
+# Reproject back to WGS84 for valid longitude and latitude
+joined_data <- st_transform(joined_data, crs = 4326)
+
 # Track school zones without cameras
 joined_data <- joined_data %>%
   mutate(no_camera_in_radius = ifelse(is.na(sign_id), TRUE, FALSE))  # Use `sign_id` to track camera presence
@@ -104,15 +112,19 @@ final_joined_data <- joined_data %>%
   inner_join(speed_counts_data_cleaned, by = "sign_id", relationship = "many-to-many")  # Specify many-to-many relationship
 
 #### Step 9: Filter Dates and Select Specific Columns ####
-
-# Filter the data
 final_filtered_data <- final_joined_data %>%
   filter(as.Date(end_date) >= as.Date("2023-01-01"),          # Keep rows with end_date in 2023 or later
          !grepl("^2024", as.character(start_date))) %>%       # Exclude rows with start_date starting with 2024
   select(sign_id, X_id, longitude, latitude,                 # Select relevant columns
-         speed_limit, volume, speed_bin, no_camera_in_radius)
+         speed_limit, volume, speed_bin, no_camera_in_radius) %>%
+  filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
+
+problem_rows <- final_filtered_data %>%
+  filter(
+    longitude <= -80 | longitude >= -78 |
+      latitude <= 43 | latitude >= 44
+  )
+print(problem_rows)
 
 #### Step 10: Save Results ####
-
-# Save the final filtered data as a Parquet file
 arrow::write_parquet(final_filtered_data, "data/analysis_data/analysis_data.parquet")

### 2024-12-02 scripts/03-test_analysis_data.R

@@ -3,25 +3,53 @@
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
-# Pre-requisites: Need to have simulated data
+# Pre-requisites: Run 03-clean_data.R script
 # Any other information needed? None.
-
 
 #### Workspace setup ####
 library(tidyverse)
-
+library(arrow)
+library(testthat)
 
 #### Test data ####
-data <- read_csv("data/raw_data/simulated.csv")
 
-# Test for negative numbers
-data$number_of_marriage |> min() <= 0
+# Load the cleaned dataset
+final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")
 
-# Test for NAs
-all(is.na(data$number_of_marriage))
+# Define tests
+test_that("Check for valid date ranges", {
+  # Ensure `end_date` is in the correct range
+  expect_true(all(as.Date(final_filtered_data$end_date) >= as.Date("2023-01-01")))
+  # Ensure `start_date` does not contain 2024 dates
+  expect_false(any(grepl("^2024", as.character(final_filtered_data$start_date))))
+})
+
+test_that("Check for missing values", {
+  # Ensure no missing values in critical columns
+  critical_cols <- c("sign_id", "X_id", "longitude", "latitude", "speed_limit", "volume", "speed_bin", "no_camera_in_radius")
+  expect_true(all(sapply(critical_cols, function(col) !any(is.na(final_filtered_data[[col]])))))
+})
+
+test_that("Check for coordinate validity", {
+  # Ensure longitude and latitude are not NA
+  expect_true(all(!is.na(final_filtered_data$longitude)), info = "Longitude contains NA values.")
+  expect_true(all(!is.na(final_filtered_data$latitude)), info = "Latitude contains NA values.")
+  
+  # Ensure longitude and latitude are within Toronto's approximate bounds
+  expect_true(all(final_filtered_data$longitude > -80 & final_filtered_data$longitude < -78), 
+              info = "Longitude values are out of bounds.")
+  expect_true(all(final_filtered_data$latitude > 43 & final_filtered_data$latitude < 44), 
+              info = "Latitude values are out of bounds.")
+})
 
 
+test_that("Check for valid speed limits and volumes", {
+  # Speed limits and volumes should be non-negative
+  expect_true(all(final_filtered_data$speed_limit >= 0))
[…]

### 2024-12-03 paper/paper.qmd
update bib from docs

@@
 bibliography: references.bib
+toc: true
+fig_caption: yes
 ---
@@
 library(tidyverse)
-
-data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
 
@@
 #| echo: false
-
-data |> 
-  ggplot(aes(x = date, y = marriage_licenses)) +
-  geom_point()
 
@@
 
-# Additional data details
+If you're incorporating feedback to diversify question types and address previous gaps, here’s an improved version of your appendix for Toronto's speeding analysis with a reasonable budget.
+I'll also ensure it reflects insights from your earlier examples and feedback while incorporating poll aggregation and methodology details.
+
+---
+
+# Idealized Survey & Methodology - $500K Budget {#sec-ideal-surv}
+
+## Overview
+This appendix outlines a survey methodology designed to enrich the analysis of excessive speeding near Toronto school zones.
+With a $500K budget, this approach aims to gather nuanced data from drivers, pedestrians, and community members about speeding behaviors, awareness of enforcement measures, and perceived safety.
+By integrating observational data with survey insights, this methodology enables robust analysis and informed policy recommendations.
+
+---
+
+## Budget Justification
+# Overview
+
+A $500,000 budget is proposed for analyzing speeding trends in Toronto school zones.
+Based on similar city-wide traffic safety initiatives, such as Vision Zero and speed enforcement studies, this budget is allocated to account for:
+
+Comprehensive sampling: Targeting diverse demographics and road users.
+
+Advanced recruitment and incentives: Ensuring adequate participation.
+
+High-quality data integration: Focused on cleaning, validation, and geospatial analysis.
+
+# Proportionality
+
+Vision Zero allocates $250,000 to $1,000,000 for focused safety measures annually, making $500,000 appropriate for a targeted study [@citytoronto2024].
+
+# Survey Costs
+
+Recruiting diverse participants typically costs $200,000 to $300,000, ensuring robust and representative data [@pew2023].
+
+---
+
+## Sampling Approach
+### **Target Groups**
+1. **Drivers:**  
+   - Key demographic: Ages 18-40, with emphasis on young male drivers.
+   - Behavioral patterns: Frequent commuters in school zones.
+
+2. **Pedestrians:**  
+   - Parents, school staff, and local residents.
+
+3. **Community Advocates:**  
+   - Stakeholders involved in traffic safety initiatives.
+
+### **Sampling Method**
+**Multi-Stage Sampling:**
+1. **Stratification by Wards:**  
+   - Ensure geographic coverage across all Toronto wards with school zones.  
+2. **Cluster Sampling:**  
+   - Focus on areas near speed cameras and school safety zones.  
+3. **Random Sampling:**  
+   - Within clusters, randomly select participants for unbiased representation.
+
+**Sample Size Goal:**  
+- **3,000 respondents**: Sufficient to identify trends at the ward level while maintaining a margin of error below ±2%.
+
+---
+
+## Recruitment Strategy
+### **Primary Channels**
+1. **Online Ads:**  
+   - Platforms: TikTok, Instagram, and YouTube.
+   - Geo-targeted to school zones and high-traffic areas.
+   - Budget Allocation: $150,000.
+
+2. **On-the-Ground Recruitment:**  
+   - Flyers distributed in school zones, community centers, and traffic courts.
+   - Budget Allocation: $50,000.
+
+3. **Partnerships with Apps:**  
+   - Collaboration with navigation tools like Waze to deliver in-app survey prompts.
+   - Budget Allocation: $100,000.
+
+4. **Incentives:**  
+   - Monetary rewards: $20 gift cards for completion, with a chance to win larger prizes.
+   - Budget Allocation: $100,000.
+
+---
+
+## Survey Design
+
+### Diversified Question Types
+To address the feedback, the survey incorporates **varied question formats**:
+
+1. **Multiple Choice:**  
+   - Example: "What factors most influence your speed in school zones?"
+     - Time pressure.
+     - Peer behavior.
+     - Lack of awareness of speed limits.
+     - Others.
+
+2. **Likert Scales:**  
+   - Example: "Rate your agreement with: 'Speed cameras make me feel safer in school zones.'"
+     - Strongly disagree to strongly agree.
+
+3. **Open-Ended:**  
+   - Example: "What other measures would help reduce speeding in school zones?"
+
+4. **Ranking Questions:**  
+   - Example: "Rank the following in order of their effectiveness at reducing speeding: speed cameras, visible signage, increased fines, police presence."
+
+5. **Interactive Scenario Questions:**  
+   - Example: "If you saw a speed camera sign but no visible camera, how likely are you to still speed? (Very unlikely to very likely)."
+
+---
+
+### Survey Flow
+1. **Introduction and Consent:**  
+   - Explain the purpose, time required (5-7 minutes), and confidentiality.
+
+2. **Demographics:**  
+   - Age, gender, driving habits, and proximity to school zones.
+
+3. **Behavioral Insights:**  
+   - Driving patterns, awareness of speed limits, and interaction with enforcement measures.
+
+4. **Perceived Effectiveness:**  
+   - Opinions on speed cameras, signage, and enforcement presence.
+
+5. **Safety Concerns:**  
+   - Perceptions of safety in school zones and suggested improvements.
+
+---
+
+## Data Validation
+1. **Logic Checks:**  
+   - Identify inconsistencies (e.g., reporting no awareness of cameras but frequent encounters).
+
+2. **Cross-Validation:**  
+   - Compare survey responses with observational data (e.g., traffic camera footage and speed counts).
+
+3. **Geospatial Matching:**  
+   - Integrate survey responses with geospatial data from speed signs and cameras for location-specific analysis.
+
+---
+
+## Poll Aggregation and Modeling
+### **Aggregation Approach**
+A **poll-of-polls method** balances survey insights with observational data:
+- Weight survey responses by:
+  - **Proximity to enforcement measures.**
+  - **Frequency of driving in school zones.**
+- Integrate responses with speed count data to build predictive models of speeding behavior.
+
+### **Modeling Framework**
+- **Dependent Variable:** Likelihood of exceeding speed limits.  
+- **Predictors:**  
+  - Awareness of enforcement measures.  
+  - Demographics (age, gender).  
+  - Self-reported speeding behavior.  
+  - Proximity to cameras and speed signs.
+
+### **Diagnostics**
+- Residual analysis and goodness-of-fit tests to validate model performance.
+- Q-Q plots and distribution checks for robustness.
+
+---
+
+## Budget Breakdown
+
+| **Expense**                     | **Cost Estimate (CAD)**  |
+|----------------------------------|--------------------------|
+| Social Media Recruitment        | $150,000                 |
+| In-App Ads and Recruitment      | $100,000                 |
+| On-the-Ground Recruitment       | $50,000                  |
+| Participant Incentives          | $100,000                 |
+| Data Cleaning and Validation    | $75,000                  |
+| Survey Design and Implementation| $25,000                  |
+| **Total**                       | **$500,000**             |
+
+---
+
+## Appendix Content
+1. **Survey Link:**  
+   [Insert Google Forms Link Here].
+
+2. **Copy of Survey Questions:**  
+   - Include all questions, organized by section.
+
+3. **Model Diagnostics:**  
+   - Residual plots, fit comparisons, and distribution checks.
+
+---
+
+This improved version addresses the feedback, diversifies question types, and integrates poll aggregation and modeling. Let me know if you’d like refinements!
 

### 2024-12-03 scripts/02-data_cleaning.R
preamble

@@ -3,7 +3,7 @@
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
-# Pre-requisites: Need to have downloaded the data
+# Pre-requisites: Need to have downloaded the data in 01-dornload_data.R script
 # Any other information needed? None.
 
 #### Workspace setup ####

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -3,7 +3,7 @@
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
-# Pre-requisites: Run 03-clean_data.R script
+# Pre-requisites: Run 02-data_cleaning.R script
 # Any other information needed? None.
 
 #### Workspace setup ####

### 2024-12-03 scripts/00-simulate_data.R
minor updates+ add sim data & some sim tests

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Simulates data
+# Purpose: Simulates data for speeding analysis
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
@@ -8,31 +8,162 @@
 
 #### Workspace setup ####
 library(tidyverse)
-
+library(ggplot2)
 
 #### Simulate data ####
 set.seed(777)
 
-# Define the start and end date
-start_date <- as.Date("2018-01-01")
-end_date <- as.Date("2023-12-31")
+# Define variables
+wards <- 1:25  # Toronto ward numbers
+speed_bins <- c("[55,60)", "[10,15)", "[35,40)", "[30,35)", "[15,20)", "[85,90)", "[25,30)", "[60,65)", "[40,45)", 
+                "[5,10)", "[100,)", "[50,55)", "[45,50)", "[20,25)", "[75,80)", "[80,85)", "[90,95)", "[95,100)", 
+                "[70,75)", "[65,70)")
 
-# Set the number of random dates you want to generate
-number_of_dates <- 100
+# Custom probabilities for speed bins
+speed_bin_probabilities <- c(
+  "[55,60)" = 0.12,
+  "[10,15)" = 0.10,
+  "[35,40)" = 0.10,
+  "[30,35)" = 0.10,
+  "[15,20)" = 0.10,
+  "[85,90)" = 0.01,  # Extremely unlikely
+  "[25,30)" = 0.10,
+  "[60,65)" = 0.12,
+  "[40,45)" = 0.12,
+  "[5,10)" = 0.08,
+  "[100,)" = 0.005,  # Rare extreme speeds
+  "[50,55)" = 0.12,
+  "[45,50)" = 0.12,
+  "[20,25)" = 0.10,
+  "[75,80)" = 0.03,  # Less likely
+  "[80,85)" = 0.02,  # Less likely
+  "[90,95)" = 0.005, # Extremely unlikely
+  "[95,100)" = 0.005, # Extremely unlikely
+  "[70,75)" = 0.05,  # Less likely
+  "[65,70)" = 0.08
+)
 
-data <-
-  tibble(
-    dates = as.Date(
-      runif(
-        n = number_of_dates,
-        min = as.numeric(start_date),
-        max = as.numeric(end_date)
-      ),
-      origin = "1970-01-01"
-    ),
-    number_of_marriage = rpois(n = number_of_dates, lambda = 15)
[…]

### 2024-12-03 scripts/02-data_cleaning.R

@@ -116,7 +116,7 @@ final_filtered_data <- final_joined_data %>%
   filter(as.Date(end_date) >= as.Date("2023-01-01"),          # Keep rows with end_date in 2023 or later
          !grepl("^2024", as.character(start_date))) %>%       # Exclude rows with start_date starting with 2024
   select(sign_id, X_id, longitude, latitude,                 # Select relevant columns
-         speed_limit, volume, speed_bin, no_camera_in_radius) %>%
+         speed_limit, volume, speed_bin, ward_no, no_camera_in_radius) %>%
   filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
 
 problem_rows <- final_filtered_data %>%

### 2024-12-03 scripts/06_test_simulated_data.R

@@ -1,0 +1,42 @@
+#### Preamble ####
+# Purpose: Tests the structure and validity of the Simulated Data
+# Author: Karen Riani
+# Date: 1 December 2024
+# Contact: karen.riani@mail.utoronto.ca
+# Pre-requisites: Run 00-simulate_data.R 
+# Any other information needed? None.
+
+#### Workspace setup ####
+
+library(testthat)
+
+# Test for duplicates
+test_that("Ensure no duplicate rows", {
+  expect_true(nrow(data.frame(final_filtered_data)) == nrow(distinct(final_filtered_data)))
+})
+
+# Test for valid longitude and latitude ranges
+test_that("Check for coordinate validity", {
+  expect_true(all(final_filtered_data$longitude > -80 & final_filtered_data$longitude < -78), 
+              info = "Longitude values are out of bounds.")
+  expect_true(all(final_filtered_data$latitude > 43 & final_filtered_data$latitude < 44), 
+              info = "Latitude values are out of bounds.")
+})
+
+# Test for valid speed limits
+test_that("Check for speed limit validity", {
+  expect_true(all(final_filtered_data$speed_limit %in% c(30, 40, 50, 60)), 
+              info = "Speed limit contains invalid values.")
+})
+
+# Test for valid volumes
+test_that("Check for valid volume values", {
+  expect_true(all(final_filtered_data$volume > 0), info = "Volume contains zero or negative values.")
+})
+
+# Test for consistency in speed_bin
+test_that("Check for valid speed_bin ranges", {
+  valid_bins <- c("[0,10)", "[10,20)", "[20,30)", "[30,40)", "[40,50)", "[50,60)", "[60,70)", "[70,80)", "[80,90)", "[90,100)")
+  expect_true(all(final_filtered_data$speed_bin %in% valid_bins), 
+              info = "Speed_bin contains invalid ranges.")
+})

### 2024-12-03 scripts/00-simulate_data.R
change path of sim

@@ -166,4 +166,4 @@ ggplot(ward_speeding_volume, aes(x = factor(ward_no), y = total_speeding_volume,
 
 #### Save Data ####
 
-write_csv(simulated_data, file = "data/raw_data/simulated.csv")
+write_csv(simulated_data, file = "data/simulated_data/simulated.csv")

### 2024-12-03 scripts/00-simulate_data.R
add detail

@@ -21,45 +21,47 @@ speed_bins <- c("[55,60)", "[10,15)", "[35,40)", "[30,35)", "[15,20)", "[85,90)"
 
 # Custom probabilities for speed bins
 speed_bin_probabilities <- c(
-  "[55,60)" = 0.12,
+  "[55,60)" = 0.15,
   "[10,15)" = 0.10,
   "[35,40)" = 0.10,
   "[30,35)" = 0.10,
   "[15,20)" = 0.10,
-  "[85,90)" = 0.01,  # Extremely unlikely
-  "[25,30)" = 0.10,
+  "[85,90)" = 0.002,   # Rare high-speed bin, ≤ 0.3%
+  "[25,30)" = 0.12,
   "[60,65)" = 0.12,
   "[40,45)" = 0.12,
   "[5,10)" = 0.08,
-  "[100,)" = 0.005,  # Rare extreme speeds
+  "[100,)" = 0.001,    # Extremely rare, ≤ 0.1%
   "[50,55)" = 0.12,
   "[45,50)" = 0.12,
   "[20,25)" = 0.10,
-  "[75,80)" = 0.03,  # Less likely
-  "[80,85)" = 0.02,  # Less likely
-  "[90,95)" = 0.005, # Extremely unlikely
-  "[95,100)" = 0.005, # Extremely unlikely
-  "[70,75)" = 0.05,  # Less likely
+  "[75,80)" = 0.02,    # Moderately rare
+  "[80,85)" = 0.003,   # Moderately rare, ≤ 0.3%
+  "[90,95)" = 0.001,   # Very rare
+  "[95,100)" = 0.0005, # Very rare
+  "[70,75)" = 0.005,
   "[65,70)" = 0.08
 )
 
 # Normalize probabilities (to ensure they sum to 1)
 speed_bin_probabilities <- speed_bin_probabilities / sum(speed_bin_probabilities)
-ward_probabilities <- runif(length(wards), 0.1, 0.3)  # Probability of high-speed bins by ward
+
+# Generate ward probabilities using a Beta distribution skewed towards lower values
+ward_probabilities <- rbeta(length(wards), shape1 = 2, shape2 = 8) * 0.2  # Scale to max 0.2
 
 # Number of observations
 num_observations <- 5000  
 
 # Simulate data
 simulated_data <- tibble(
-  sign_id = sample(unique(final_filtered_data$sign_id), num_observations, replace = TRUE),  # Random sign IDs
[…]

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -53,3 +53,26 @@ test_that("Verify `no_camera_in_radius` logic", {
   # Ensure `no_camera_in_radius` is TRUE only when `sign_id` is NA
   expect_true(all(final_filtered_data$no_camera_in_radius == is.na(final_filtered_data$sign_id)))
 })
+
+test_that("Check for proportion of extreme speed bins", {
+  extreme_speeds <- final_filtered_data %>%
+    filter(speed_bin == "[100,)")
+  expect_true(nrow(extreme_speeds) / nrow(final_filtered_data) <= 0.01, 
+              info = "Extreme speed bins exceed 1% of the data.")
+  
+  high_speed_bins <- c("[85,90)", "[90,95)", "[95,100)")
+  high_speed_proportion <- final_filtered_data %>%
+    filter(speed_bin %in% high_speed_bins) %>%
+    nrow() / nrow(final_filtered_data)
+  expect_true(high_speed_proportion < 0.03, 
+              info = "High-speed bins exceed 3% of the data.")
+})
+
+test_that("Ensure speed bins are valid and present", {
+  valid_speed_bins <- c("[55,60)", "[10,15)", "[35,40)", "[30,35)", "[15,20)", 
+                        "[85,90)", "[25,30)", "[60,65)", "[40,45)", "[5,10)", 
+                        "[100,)", "[50,55)", "[45,50)", "[20,25)", "[75,80)", 
+                        "[80,85)", "[90,95)", "[95,100)", "[70,75)", "[65,70)")
+  expect_true(all(final_filtered_data$speed_bin %in% valid_speed_bins), 
+              info = "Invalid or missing speed_bin values found.")
+})

### 2024-12-03 scripts/05-exploratory_analysis.R

@@ -1,19 +1,75 @@
 #### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Purpose:  Exploratory Data Analysis for Cleaned Dataset
+# Author: Karen Riani
+# Date: 3 December 2024
+# Contact: karen.riani@mail.utoronto.ca
+# Pre-requisites: Run 02-data_cleaning.R script
+# Any other information needed? None.
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(arrow)
+library(testthat)
 
-#### Load data ####
-# [...ADD CODE HERE...]
+#### Test data ####
 
+# Load the cleaned dataset
+final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")
+
+
+
+
+
+
+
+
+
+# Calculate statistics for speeds over 100 km/h
+speeding_stats <- raw_data %>%
+  summarise(
+    # Number of rows with any occurrence of speeding over 100
+    rows_with_speeding = sum(spd_100_and_above > 0, na.rm = TRUE),
+    
+    # Total occurrences of speeding over 100 summed across all rows
+    total_speeding_count = sum(spd_100_and_above, na.rm = TRUE)
+  )
+print(speeding_stats)
+
+
+view by ward... need ward data though
+
+]
+library(leaflet)
+
+# Create a Leaflet map
+leaflet(data = filtered_data) %>%
+  addTiles() %>%
+  addCircles(
+    lng = ~longitude, lat = ~latitude, weight = 1,
+    radius = ~volume * 10,  # Adjust radius based on traffic volume
+    color = "red", fillColor = "red", fillOpacity = 0.5
+  ) %>%
+  addLegend(
+    "bottomright", pal = colorNumeric("YlOrRd", domain = NULL), 
+    values = ~volume, title = "Traffic Volume"
+  ) %>%
+  setView(lng = -79.3832, lat = 43.6532, zoom = 12)  # Centered on Toronto
+
+
+speeding_data <- final_filtered_data %>%
[…]

### 2024-12-03 scripts/06_test_simulated_data.R

@@ -9,34 +9,85 @@
 #### Workspace setup ####
 
 library(testthat)
+library(tidyverse)
 
-# Test for duplicates
-test_that("Ensure no duplicate rows", {
-  expect_true(nrow(data.frame(final_filtered_data)) == nrow(distinct(final_filtered_data)))
-})
+simulated_data <- read_csv("data/simulated_data/simulated.csv")
 
-# Test for valid longitude and latitude ranges
-test_that("Check for coordinate validity", {
-  expect_true(all(final_filtered_data$longitude > -80 & final_filtered_data$longitude < -78), 
-              info = "Longitude values are out of bounds.")
-  expect_true(all(final_filtered_data$latitude > 43 & final_filtered_data$latitude < 44), 
-              info = "Latitude values are out of bounds.")
-})
+# Test if the data was successfully loaded
+if (exists("simulated_data")) {
+  message("Test Passed: The dataset was successfully loaded.")
+} else {
+  stop("Test Failed: The dataset could not be loaded.")
+}
 
-# Test for valid speed limits
-test_that("Check for speed limit validity", {
-  expect_true(all(final_filtered_data$speed_limit %in% c(30, 40, 50, 60)), 
-              info = "Speed limit contains invalid values.")
-})
+#### Test Simulated Speeding Data ####
 
-# Test for valid volumes
-test_that("Check for valid volume values", {
-  expect_true(all(final_filtered_data$volume > 0), info = "Volume contains zero or negative values.")
-})
+# Expected values
+expected_speed_bins <- c("[55,60)", "[10,15)", "[35,40)", "[30,35)", "[15,20)", "[85,90)", "[25,30)", "[60,65)", 
+                         "[40,45)", "[5,10)", "[100,)", "[50,55)", "[45,50)", "[20,25)", "[75,80)", "[80,85)", 
+                         "[90,95)", "[95,100)", "[70,75)", "[65,70)")
+expected_wards <- 1:25  # Toronto wards
+expected_speed_limits <- c(30, 40, 50, 60)
 
-# Test for consistency in speed_bin
-test_that("Check for valid speed_bin ranges", {
-  valid_bins <- c("[0,10)", "[10,20)", "[20,30)", "[30,40)", "[40,50)", "[50,60)", "[60,70)", "[70,80)", "[80,90)", "[90,100)")
[…]

### 2024-12-03 scripts/02-data_cleaning.R
add speed over amt col + updates

@@ -119,12 +119,14 @@ final_filtered_data <- final_joined_data %>%
          speed_limit, volume, speed_bin, ward_no, no_camera_in_radius) %>%
   filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
 
-problem_rows <- final_filtered_data %>%
-  filter(
-    longitude <= -80 | longitude >= -78 |
-      latitude <= 43 | latitude >= 44
+# Add a column for the lower end of the speed bin and calculate the amount over the speed limit
+final_filtered_data <- final_filtered_data %>%
+  mutate(
+    # Extract the lower bound of the speed bin
+    speed_bin_lower = as.numeric(gsub("\\[|,.*", "", speed_bin)),
+    # Calculate the amount over the speed limit
+    over_speed_limit = pmax(speed_bin_lower - speed_limit, 0)  # Use pmax to ensure no negative values
   )
-print(problem_rows)
 
 #### Step 10: Save Results ####
 arrow::write_parquet(final_filtered_data, "data/analysis_data/analysis_data.parquet")

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -76,3 +76,23 @@ test_that("Ensure speed bins are valid and present", {
   expect_true(all(final_filtered_data$speed_bin %in% valid_speed_bins), 
               info = "Invalid or missing speed_bin values found.")
 })
+
+test_that("Check for correctness of over_speed_limit calculation", {
+  # Check if over_speed_limit is non-negative
+  expect_true(all(final_filtered_data$over_speed_limit >= 0), 
+              info = "over_speed_limit contains negative values.")
+  
+  # Check if over_speed_limit matches lower speed_bin - speed_limit for sample rows
+  sample_data <- final_filtered_data %>%
+    slice_sample(n = 10)  # Take 10 random rows for testing
+  
+  sample_data <- sample_data %>%
+    mutate(
+      expected_over_speed_limit = pmax(as.numeric(gsub("\\[|,.*", "", speed_bin)) - speed_limit, 0)
+    )
+  
+  # Ensure calculated and expected values match
+  expect_true(all(sample_data$over_speed_limit == sample_data$expected_over_speed_limit),
+              info = "over_speed_limit calculation does not match expected values.")
+})
+

### 2024-12-03 scripts/05-exploratory_analysis.R

@@ -10,52 +10,14 @@
 library(tidyverse)
 library(arrow)
 library(testthat)
+library(corrplot)
 
 #### Test data ####
 
 # Load the cleaned dataset
 final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")
 
-
-
-
-
-
-
-
-
-# Calculate statistics for speeds over 100 km/h
-speeding_stats <- raw_data %>%
-  summarise(
-    # Number of rows with any occurrence of speeding over 100
-    rows_with_speeding = sum(spd_100_and_above > 0, na.rm = TRUE),
-    
-    # Total occurrences of speeding over 100 summed across all rows
-    total_speeding_count = sum(spd_100_and_above, na.rm = TRUE)
-  )
-print(speeding_stats)
-
-
-view by ward... need ward data though
-
-]
-library(leaflet)
-
-# Create a Leaflet map
-leaflet(data = filtered_data) %>%
-  addTiles() %>%
-  addCircles(
-    lng = ~longitude, lat = ~latitude, weight = 1,
-    radius = ~volume * 10,  # Adjust radius based on traffic volume
-    color = "red", fillColor = "red", fillOpacity = 0.5
-  ) %>%
-  addLegend(
-    "bottomright", pal = colorNumeric("YlOrRd", domain = NULL), 
-    values = ~volume, title = "Traffic Volume"
-  ) %>%
-  setView(lng = -79.3832, lat = 43.6532, zoom = 12)  # Centered on Toronto
-
-
+# Plot proportion of general speeding across wards
 speeding_data <- final_filtered_data %>%
   mutate(speeding_over_limit = as.numeric(gsub("[^0-9]", "", gsub("\\)|\\]", "", speed_bin))) - speed_limit) %>%
   filter(speeding_over_limit > 0) %>%
@@ -72,4 +34,101 @@ ggplot(speeding_data, aes(x = reorder(ward_no, proportion_speeding), y = proport
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
+# Add derived variables for speeding and extreme speeding
+final_filtered_data <- final_filtered_data %>%
+  mutate(
+    speed_bin_numeric = as.numeric(gsub("\\D", "", speed_bin)) + 5,  # Approximate midpoint
+    speeding = ifelse(speed_bin_numeric > speed_limit, 1, 0),  # Indicator for speeding
[…]

### 2024-12-03 scripts/02-data_cleaning.R
fix join issue

@@ -107,9 +107,9 @@ speed_counts_data_cleaned <- speed_counts_data_raw %>%
   ungroup()
 
 #### Step 8: Join Speed Counts with Joined Camera and Speed data #### 
-# Perform an inner join to ensure no NA values in `sign_id`
+# Perform a left join to include all rows from `joined_data`, even if `sign_id` is NA
 final_joined_data <- joined_data %>%
-  inner_join(speed_counts_data_cleaned, by = "sign_id", relationship = "many-to-many")  # Specify many-to-many relationship
+  left_join(speed_counts_data_cleaned, by = "sign_id", relationship = "many-to-many")  # Specify many-to-many relationship
 
 #### Step 9: Filter Dates and Select Specific Columns ####
 final_filtered_data <- final_joined_data %>%

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -26,7 +26,7 @@ test_that("Check for valid date ranges", {
 
 test_that("Check for missing values", {
   # Ensure no missing values in critical columns
-  critical_cols <- c("sign_id", "X_id", "longitude", "latitude", "speed_limit", "volume", "speed_bin", "no_camera_in_radius")
+  critical_cols <- c("longitude", "latitude", "speed_limit", "volume", "speed_bin", "no_camera_in_radius")
   expect_true(all(sapply(critical_cols, function(col) !any(is.na(final_filtered_data[[col]])))))
 })
 

### 2024-12-03 scripts/04-model.R
little updates

@@ -1,30 +1,20 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
+# Purpose:  Checking and Fitting Models and Making Predictions
 # Author: Karen Riani
-# Date: 1 December 2024
+# Date: 3 December 2024
 # Contact: karen.riani@mail.utoronto.ca
-# Pre-requisites: None
-# Any other information needed? None
-
+# Pre-requisites: Run 02-data_cleaning.R script
+# Any other information needed? None.
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
+library(arrow)
+library(testthat)
 
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+#### Test data ####
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+# Load the cleaned dataset
+final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")
 
 
 #### Save model ####

### 2024-12-03 scripts/02-data_cleaning.R
add checks and summaries

@@ -119,6 +119,13 @@ final_filtered_data <- final_joined_data %>%
          speed_limit, volume, speed_bin, ward_no, no_camera_in_radius) %>%
   filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
 
+critical_cols <- c("speed_limit", "volume", "speed_bin", "no_camera_in_radius")
+
+# Remove rows with NA in critical columns
+final_joined_data <- final_joined_data %>%
+  filter(if_any(all_of(critical_cols), ~ !is.na(.)))
+
+
 # Add a column for the lower end of the speed bin and calculate the amount over the speed limit
 final_filtered_data <- final_filtered_data %>%
   mutate(

### 2024-12-03 scripts/05-exploratory_analysis.R

@@ -57,7 +57,8 @@ summary_stats <- final_filtered_data %>%
 # Display summary
 summary_stats
 
-# Model Summary
+# General Summary
+summary(final_filtered_data)
 
 # Summarize speeding and extreme speeding by ward
 ward_speeding_summary <- final_filtered_data %>%

### 2024-12-03 scripts/02-data_cleaning.R
fix logic issue, remove no cam in radius

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Cleans the raw marriage data into an analysis dataset
+# Purpose: Cleans the raw speed and camera data into an analysis dataset
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
@@ -89,10 +89,6 @@ if (is.na(st_crs(joined_data))) {
 # Reproject back to WGS84 for valid longitude and latitude
 joined_data <- st_transform(joined_data, crs = 4326)
 
-# Track school zones without cameras
-joined_data <- joined_data %>%
-  mutate(no_camera_in_radius = ifelse(is.na(sign_id), TRUE, FALSE))  # Use `sign_id` to track camera presence
-
 #### Step 6: Convert geometry back into longitude and latitude columns #### 
 joined_data <- joined_data %>%
   mutate(longitude = st_coordinates(.)[, 1],  # Extract X (longitude)
@@ -107,33 +103,26 @@ speed_counts_data_cleaned <- speed_counts_data_raw %>%
   ungroup()
 
 #### Step 8: Join Speed Counts with Joined Camera and Speed data #### 
-# Perform a left join to include all rows from `joined_data`, even if `sign_id` is NA
+# Perform an inner join to ensure no NA values in `sign_id`
 final_joined_data <- joined_data %>%
-  left_join(speed_counts_data_cleaned, by = "sign_id", relationship = "many-to-many")  # Specify many-to-many relationship
+  inner_join(speed_counts_data_cleaned, by = "sign_id", relationship = "many-to-many")  # Specify many-to-many relationship
 
 #### Step 9: Filter Dates and Select Specific Columns ####
 final_filtered_data <- final_joined_data %>%
   filter(as.Date(end_date) >= as.Date("2023-01-01"),          # Keep rows with end_date in 2023 or later
          !grepl("^2024", as.character(start_date))) %>%       # Exclude rows with start_date starting with 2024
   select(sign_id, X_id, longitude, latitude,                 # Select relevant columns
-         speed_limit, volume, speed_bin, ward_no, no_camera_in_radius) %>%
+         speed_limit, volume, speed_bin, ward_no) %>%
   filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
 
-critical_cols <- c("speed_limit", "volume", "speed_bin", "no_camera_in_radius")
-
-# Remove rows with NA in critical columns
[…]

### 2024-12-03 scripts/02-data_cleaning.R
camera issue updates + minot fixes

@@ -4,7 +4,7 @@
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
 # Pre-requisites: Need to have downloaded the data in 01-dornload_data.R script
-# Any other information needed? None.
+# Any other information needed? Camera-related variables, such as camera density or proximity, were not included in the main data-cleaning pipeline as they were found to be highly correlated with ward_no due to the spatial join process. These variables are explored separately in the model and exploratory analysis scripts to assess their utility
 
 #### Workspace setup ####
 library(tidyverse)
@@ -115,7 +115,7 @@ final_filtered_data <- final_joined_data %>%
          speed_limit, volume, speed_bin, ward_no) %>%
   filter(!is.na(longitude) & !is.na(latitude))  # Remove rows with missing coordinates
 
-  # Add a column for the lower end of the speed bin and calculate the amount over the speed limit
+# Add a column for the lower end of the speed bin and calculate the amount over the speed limit
 final_filtered_data <- final_filtered_data %>%
       mutate(
         # Extract the lower bound of the speed bin

### 2024-12-03 scripts/05-exploratory_analysis.R

@@ -14,7 +14,7 @@ library(corrplot)
 
 #### Test data ####
 
-# Load the cleaned dataset
+# Load the datasets
 final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")
 
 # Plot proportion of general speeding across wards
@@ -82,7 +82,6 @@ ggplot(ward_speeding_summary, aes(x = factor(ward_no), y = total_speeding, fill 
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
                                      
 
-
 # Interactive map of speeding incidents
 leaflet(final_filtered_data %>% filter(speeding == 1)) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
@@ -133,3 +132,24 @@ ggplot(heatmap_data, aes(x = factor(ward_no), y = over_speed_limit, fill = total
     axis.text.x = element_text(angle = 45, hjust = 1),
     panel.grid = element_blank()
   )
+
+### Explore camera density per ward
+camera_density <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(cameras_nearby = n_distinct(X_id), .groups = "drop")
+
+ward_speed_analysis <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(
+    avg_speed_over_limit = mean(over_speed_limit, na.rm = TRUE),
+    cameras_nearby = n_distinct(X_id),
+    .groups = "drop"
+  )
+
+ggplot(ward_speed_analysis, aes(x = cameras_nearby, y = avg_speed_over_limit)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(title = "Camera Density vs. Speeding Behavior",
+       x = "Number of Cameras Nearby",
+       y = "Average Speed Over Limit")
+

### 2024-12-04 scripts/04-model.R
finalize models

@@ -8,19 +8,130 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(tidymodels)
 library(arrow)
-library(testthat)
-
-#### Test data ####
+library(rstanarm)
+library(pROC)
 
 # Load the cleaned dataset
 final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")
 
+# Convert `ward_no` to a factor for categorical analysis
+final_filtered_data <- final_filtered_data %>%
+  mutate(ward_no = factor(ward_no))
 
-#### Save model ####
+
+#### Test data ####
+
+### Checking for collinearity in camera columns
+# Calculate camera density
+camera_density <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(cameras_nearby = n_distinct(X_id), .groups = "drop")
+
+# Add to the final_filtered_data
+cam_filtered_data <- final_filtered_data %>%
+  left_join(camera_density, by = "ward_no")
+
+model_camera_density <- lm(
+  over_speed_limit ~  speed_limit + volume + cameras_nearby + ward_no,
+  data = cam_filtered_data
+)
+
+# View Collinearity
+summary(model_camera_density)
+vif(model_camera_density)
+
+
+### Modelling without camera columns
+
+## Set seed for reproducibility and split the data 
+set.seed(777)
+data_split <- initial_split(final_filtered_data, prop = 0.80) # 80-20 split
+data_train <- training(data_split)
+data_test <- testing(data_split)
+
+# Ensure consistent levels for ward_no in training and test datasets
+common_levels <- levels(factor(data_train$ward_no))
+data_train <- data_train %>%
+  mutate(ward_no = factor(ward_no, levels = common_levels))
+
+data_test <- data_test %>%
+  mutate(ward_no = factor(ward_no, levels = common_levels))
+
+# Ensure no NA in critical columns for the test set
+data_test <- data_test %>%
+  filter(if_all(c("speed_limit", "volume", "ward_no", "X_id"), ~ !is.na(.)))
+
+#### Model 1 ####
+# Fit MLR model on the training dataset
+model_1 <- lm(
+  over_speed_limit ~ speed_limit + volume + ward_no,
+  data = data_train
+)
+
+# View model summary and compute VIF
+summary(model_1)
+vif(model_1)
+
[…]

### 2024-12-04 scripts/04-model.R
little fix

@@ -11,7 +11,6 @@ library(tidyverse)
 library(tidymodels)
 library(arrow)
 library(rstanarm)
-library(pROC)
 
 # Load the cleaned dataset
 final_filtered_data <- arrow::read_parquet("data/analysis_data/analysis_data.parquet")

### 2024-12-04 paper/paper.qmd
add data paper updates

@@
 library(tidyverse)
+library(arrow)
+library(kableExtra)
+
+final_filtered_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+
+```
+\newpage
+
+# Introduction
+
+Toronto's battle against speeding takes some wild turns—literally and figuratively.
+A speed camera notorious for issuing tens of thousands of tickets on Parkside Drive, Toronto's most ticketed street for speeding, was cut down less than 24 hours after being reinstalled [@toronto2023; @cbc2023].
+While the city struggles to curb speeding, new research by the Canadian Automobile Association (CAA) paints a troubling picture: nearly 70% of Canadians admit to speeding in residential areas, 50% routinely speed on highways, and 20% regularly push well beyond the limit [@jobbernation2023].
+
+The rise of advanced vehicle technology and the adrenaline appeal of street racing exacerbate these issues, with young male drivers (18–25 years) disproportionately represented in this trend.
+Though rare at **0.9% prevalence**, street racing poses a **five-fold increase in crash risk (OR = 5.23, p < .001)**, even after accounting for other risk factors [@simonsmorton2017].
+Despite Ontario’s strict penalties, such as **$10,000 fines** and **30-day license suspensions for excessive speeding**, enforcement alone has not fully addressed these challenges [@trafficparalegal2024].
+Recent headlines, such as a driver clocked at **102 km/h in a 50 km/h school zone** and a Toronto speed camera being destroyed less than 24 hours after reinstallation, underscore the disconnect between enforcement measures and public behavior [@citynews2024; @cbc2023].
+
+This paper focuses on understanding **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit.
+This dependent variable is calculated as the difference between the lower bound of the observed speed bin and the speed limit, ensuring a straightforward measure of speeding severity.
+By quantifying how far drivers exceed the limit, over_speed_limit allows for identifying patterns of high-risk speeding behaviour.
+Using open data from Toronto's School Safety Zone Watch Your Speed Program, Watch Your Speed Detailed Counts, and Traffic Camera datasets, including vehicle counts (volume) in each speed bin, speed limits, and ward-level identifiers, this study investigates how traffic patterns, enforcement mechanisms, and geographic factors contribute to variations in speeding behaviour across the city.
+
+To address this, the study employs multi-linear regression models to examine the relationship between excess speed and predictors such as traffic volume, speed limits, and ward characteristics.
+Results indicate that higher traffic volume reduces excess speed by **0.12 km/h per additional vehicle**, while speed limits positively influence speeding by **0.13 km/h per unit increase**.
+Wards exhibit substantial variability, with some, such as Ward 4, showing significant decreases in speeding.
+However, the impact of individual speed cameras was negligible, suggesting limitations in the data and methodology for evaluating camera-specific effects.
+
+The rest of this paper is structured as follows:  @sec-mydatasection presents the data sources and methodology, @sec-mymodel presents the forecasting model, then  @sec-myresults followed by  @sec-mydisc, which discusses the results and their implications.
+@sec-mylimits concludes with a discussion of the limitations of the study and suggestions for future research, followed by Appendix @sec-ideal-surv with a budgeted idealized survey and methodology.
+
+# Data {#sec-mydatasection}
+
+## Overview
+The data for this analysis is sourced from Open Data Toronto and comprises three datasets: School Safety Zone Watch Your Speed Program – Locations [@schoolzonelocations], School Safety Zone Watch Your Speed Program – Detailed Speed Counts [@schoolzonespeeds], and the Traffic Camera dataset  [@trafficcameras].
+These datasets provide detailed insights into speeding behavior, traffic enforcement mechanisms, and geographic characteristics within Toronto.
+All data analysis was performed using R [@citeR], employing a variety of packages to ensure robust and reproducible workflows.
+
+## Packages Used
+The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `kableExtra` [@kableExtra] and `rstanarm` [@rstanarm].
+
+## Measurement and Limitations
+
+The datasets from Open Data Toronto serve as the foundation for analyzing speeding behavior in Toronto.
+Each variable is derived from real-world phenomena and represents a specific aspect of traffic patterns or enforcement mechanisms:
+
+1.
+**Over Speed Limit**: This key dependent variable quantifies the extent to which drivers exceed posted speed limits.
+It is derived by calculating the difference between the midpoint of the observed speed bin (e.g., `[45,50)` translates to a midpoint of 47.5 km/h) and the posted speed limit.
+This method provides a nuanced measure of speeding severity, distinguishing between minor infractions and extreme violations.
+
+2.
+**No Camera in Radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations.
+If a given location fell outside the defined radius of nearby cameras, it was flagged as `TRUE`.
+This captures the absence of enforcement presence and allows for an analysis of its influence on speeding behavior.
+
+3.
+**Camera Density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward.
+This variable offers insight into how concentrated enforcement might correlate with traffic patterns or speeding violations.
+
+4.
+**Traffic Volume**: This variable directly records the number of vehicles observed at a specific location within a given speed bin and time interval.
+It is included in the datasets as raw counts, reflecting real-time traffic flow.
+
+5. **Speed Limit**: This variable is a direct measure of the posted speed regulations at specific locations. Data entries were validated against known regulations to ensure accuracy.
+
+6.
+**Ward Number**: A categorical representation of Toronto’s geographic divisions, this variable contextualizes speeding behavior within specific zones.
+Each ward entry corresponds to a specific administrative boundary, linking observed behaviors to broader urban planning and enforcement efforts.
+
+### Data Transformation and Limitations
+The process of moving from real-world observations to dataset entries involves several key transformations:
+
+- **Spatial Joins**: Camera locations (latitude and longitude) were mapped against speed observation points to identify enforcement presence.
+
+- **Calculated Variables**: Metrics like over speed limit and camera density were derived from raw data, enabling a more analytical exploration of speeding patterns.
+
+Despite these transformations, limitations exist.
+Speed bin midpoints assume a uniform speed distribution within each bin, which may not reflect reality.
+Additionally, camera placements are biased toward high-traffic or high-risk areas, potentially underrepresenting quieter zones.
+These factors necessitate caution when interpreting results, as the dataset captures a snapshot rather than a complete picture of traffic behavior.
+
+## Variables of Interest
+The dataset includes several critical variables relevant to speeding behavior:
+
+- **Sign ID and Camera ID (X_id)**: Unique identifiers for speed cameras and their locations, enabling the association of speeding patterns with enforcement points.
+- **Longitude and Latitude**: Geographic coordinates capturing the location of observed speeding behavior.
+- **Speed Limit**: Posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h.
+- **Volume**: The number of vehicles observed in a given speed bin. The median volume is 13 vehicles, but counts range widely, with a maximum of 209 vehicles.
+- **Speed Bin**: Recorded ranges of observed speeds, such as `[45,50)`, providing context for speed categories.
+- **Ward Number**: Geographic zones represented by categorical identifiers.
+Ward 10 had the highest representation in the data (28.8% of observations), while Ward 19 had significantly fewer observations.
+- **Over Speed Limit**: A calculated variable representing the extent to which observed speeds exceed posted limits.
+This variable allows for a nuanced exploration of speeding severity, with a mean value of 3.91 km/h and a maximum of 70 km/h.
+
+## Descriptive Statistics
+Descriptive statistics in [@tbl-summary] highlight critical patterns within the data:
+
+- The **mean over speed limit** is 3.91 km/h, with extreme violations reaching up to 70 km/h.
+
+- The **median speed bin midpoint** is 30 km/h, while the mean traffic volume is 23.35 vehicles.
+
+- Geographic variables such as **longitude** and **latitude** center around Toronto’s urban core.
+
+```{r}
+#| label: tbl-summary
+#| fig-cap: Summary of Joined Analysis Data
+#| echo: false
+
+# Prepare summary data
+summary_data <- data.frame(
+  Variable = c("sign_id", "X_id", "longitude", "latitude", 
+               "speed_limit", "volume", "speed_bin_lower", "over_speed_limit"),
+  Min = c(23, 11, -79.58, 43.64, 30, 1, 5, 0),
+  "1st Qu." = c(667, 146, -79.40, 43.65, 40, 4, 15, 0),
+  Median = c(950, 298, -79.36, 43.65, 40, 13, 30, 0),
+  Mean = c(1105, 247.1, -79.36, 43.68, 38.24, 23.35, 30.23, 3.91),
+  "3rd Qu." = c(978, 362, -79.33, 43.71, 40, 35, 40, 5),
+  Max = c(2622, 423, -79.19, 43.79, 50, 209, 100, 70)
+)
+
+# Create a formatted table
+summary_data %>%
+  kbl(caption = "Summary Statistics for the Final Filtered Data") %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = FALSE, 
+                position = "center") %>%
+  column_spec(1, bold = TRUE, width = "10em") %>%
+  row_spec(0, bold = TRUE)
 
@@
 
-# Introduction
+# Model {#sec-mymodel}
 
-You can and should cross-reference sections and sub-sections. We use @talia, @opendatatorotno, and @rohan.
+# Results {#sec-myresults}
 
-The remainder of this paper is structured as follows. @sec-mydatasection
-
-# Data {#sec-mydatasection}
-
-Some of our data is of penguins (@fig-marriage), from @palmerpenguins.
-
-```{r}
-#| label: fig-marriage
-#| fig-cap: Bills of penguins
-#| echo: false
-
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-# Discussion
+# Discussion {#sec-mydisc}
 
@@
 
-## Weaknesses and next steps
+# Weaknesses and next steps {#sec-mylimits}
 

### 2024-12-04 scripts/02-data_cleaning.R
tiny fix

@@ -3,7 +3,7 @@
 # Author: Karen Riani
 # Date: 1 December 2024
 # Contact: karen.riani@mail.utoronto.ca
-# Pre-requisites: Need to have downloaded the data in 01-dornload_data.R script
+# Pre-requisites: Need to have downloaded the data in 01-download_data.R script
 # Any other information needed? Camera-related variables, such as camera density or proximity, were not included in the main data-cleaning pipeline as they were found to be highly correlated with ward_no due to the spatial join process. These variables are explored separately in the model and exploratory analysis scripts to assess their utility
 
 #### Workspace setup ####

### 2024-12-04 paper/paper.qmd
update from doc

@@
 
-This paper focuses on understanding **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit.
+This paper focuses on understanding the estimand **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit.
 This dependent variable is calculated as the difference between the lower bound of the observed speed bin and the speed limit, ensuring a straightforward measure of speeding severity.
@@
 
-To address this, the study employs multi-linear regression models to examine the relationship between excess speed and predictors such as traffic volume, speed limits, and ward characteristics.
+To address this, the study employs multi-linear regression models to examine the relationship between excess speed and predictors such as traffic volume, speed limits, and ward characteristics throughout 2023.
 Results indicate that higher traffic volume reduces excess speed by **0.12 km/h per additional vehicle**, while speed limits positively influence speeding by **0.13 km/h per unit increase**.
@@
 ## Overview
-The data for this analysis is sourced from Open Data Toronto and comprises three datasets: School Safety Zone Watch Your Speed Program – Locations [@schoolzonelocations], School Safety Zone Watch Your Speed Program – Detailed Speed Counts [@schoolzonespeeds], and the Traffic Camera dataset  [@trafficcameras].
+The data for this analysis is sourced from Open Data Toronto [@opendatatoronto] and comprises three datasets: School Safety Zone Watch Your Speed Program – Locations [@schoolzonelocations], School Safety Zone Watch Your Speed Program – Detailed Speed Counts in 2023 [@schoolzonespeeds], and the Traffic Camera dataset  [@trafficcameras].
 These datasets provide detailed insights into speeding behavior, traffic enforcement mechanisms, and geographic characteristics within Toronto.
-All data analysis was performed using R [@citeR], employing a variety of packages to ensure robust and reproducible workflows.
+The traffic camera and speed sign data were merged through a spacial join with a range of 500m using geographical coordinates.
+This was joined with the detailed counts of speeds through Toronto ward number (25 ward structure).
 
@@
 
-## Measurement and Limitations
+## Measurement
 
@@
 
-1.
-**Over Speed Limit**: This key dependent variable quantifies the extent to which drivers exceed posted speed limits.
+- **Over Speed Limit**: This key dependent variable quantifies the extent to which drivers exceed posted speed limits.
 It is derived by calculating the difference between the midpoint of the observed speed bin (e.g., `[45,50)` translates to a midpoint of 47.5 km/h) and the posted speed limit.
@@
 
-2.
-**No Camera in Radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations.
+- **Sign ID and Camera ID (X_id)**: These unique identifiers represent speed cameras and their locations, facilitating the mapping of speeding patterns to enforcement points.
+However, sign addresses may not always align precisely with camera placements, as discussed in the limitations section.
+
+- **Longitude and Latitude**: Geographic coordinates capturing the locations of observed speeding behaviour, allowing for spatial analysis and the identification of high-risk zones.
+
+- **Speed Limit**: The posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h. This variable provides the baseline against which speeding is measured.
+
+- **Volume**: The count of vehicles observed in a given speed bin, representing traffic density. Median volume is 13 vehicles, with counts ranging widely, up to a maximum of 209 vehicles.
+
+- **Speed Bin**: Recorded ranges of observed speeds, such as `[45,50)`, provide context for speed categories, assuming uniform speed distributions within each bin.
+
+- **Ward Number**: Geographic zones represented as categorical identifiers, reflecting administrative boundaries within Toronto.
+For instance, Ward 10 accounts for 28.8% of observations, making it the most represented ward, while Ward 19 has considerably fewer observations.
+
+- **No Camera in Radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations.
 If a given location fell outside the defined radius of nearby cameras, it was flagged as `TRUE`.
@@
 
-3.
-**Camera Density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward.
+- **Camera Density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward.
 This variable offers insight into how concentrated enforcement might correlate with traffic patterns or speeding violations.
 
-4.
-**Traffic Volume**: This variable directly records the number of vehicles observed at a specific location within a given speed bin and time interval.
-It is included in the datasets as raw counts, reflecting real-time traffic flow.
+## Data Limitations
 
-5. **Speed Limit**: This variable is a direct measure of the posted speed regulations at specific locations. Data entries were validated against known regulations to ensure accuracy.
+The datasets used in this study have notable limitations that must be acknowledged.
+**Sign addresses** in the datasets have not been verified and may represent the address of an adjacent property rather than the actual location of the sign.
+The **speed limit** column reflects the most recent operating parameter, which does not necessarily align with the current or previous speed limits for each location.
 
-6.
-**Ward Number**: A categorical representation of Toronto’s geographic divisions, this variable contextualizes speeding behavior within specific zones.
-Each ward entry corresponds to a specific administrative boundary, linking observed behaviors to broader urban planning and enforcement efforts.
+The **Traffic Cameras** dataset is highly up-to-date, with a last refresh on 2024-11-26.
+While there are no issues with accessibility or completeness, usability is constrained by unclear column names and incomplete metadata.
+Conversely, the **School Safety Zone Watch Your Speed Program – Detailed Speed Counts** dataset is over 12 months outdated, requiring reliance on direct CSV links for access.
+This reduces its applicability for analysing current traffic trends.
 
-### Data Transformation and Limitations
-The process of moving from real-world observations to dataset entries involves several key transformations:
-
-- **Spatial Joins**: Camera locations (latitude and longitude) were mapped against speed observation points to identify enforcement presence.
-
-- **Calculated Variables**: Metrics like over speed limit and camera density were derived from raw data, enabling a more analytical exploration of speeding patterns.
-
-Despite these transformations, limitations exist.
-Speed bin midpoints assume a uniform speed distribution within each bin, which may not reflect reality.
-Additionally, camera placements are biased toward high-traffic or high-risk areas, potentially underrepresenting quieter zones.
-These factors necessitate caution when interpreting results, as the dataset captures a snapshot rather than a complete picture of traffic behavior.
-
-## Variables of Interest
-The dataset includes several critical variables relevant to speeding behavior:
-
-- **Sign ID and Camera ID (X_id)**: Unique identifiers for speed cameras and their locations, enabling the association of speeding patterns with enforcement points.
-- **Longitude and Latitude**: Geographic coordinates capturing the location of observed speeding behavior.
-- **Speed Limit**: Posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h.
-- **Volume**: The number of vehicles observed in a given speed bin. The median volume is 13 vehicles, but counts range widely, with a maximum of 209 vehicles.
-- **Speed Bin**: Recorded ranges of observed speeds, such as `[45,50)`, providing context for speed categories.
-- **Ward Number**: Geographic zones represented by categorical identifiers.
-Ward 10 had the highest representation in the data (28.8% of observations), while Ward 19 had significantly fewer observations.
-- **Over Speed Limit**: A calculated variable representing the extent to which observed speeds exceed posted limits.
-This variable allows for a nuanced exploration of speeding severity, with a mean value of 3.91 km/h and a maximum of 70 km/h.
+To address gaps, **spatial joins** were employed to map camera locations to speed observation points, identifying enforcement presence, and **calculated variables**, such as over speed limit and camera density, were derived to facilitate analysis.
+However, these transformations have limitations.
+Speed bin ranges assume a uniform speed distribution, which may not capture the variability in driver behaviours.
+Furthermore, camera placements are concentrated in high-traffic or high-risk areas, potentially underrepresenting quieter zones.
+Together, these factors highlight the need for caution when interpreting results, as the datasets provide a snapshot rather than a comprehensive depiction of traffic behaviour.
 
@@
 
+```{r}
+#| label: fig-heatbar
+#| fig-cap: Number of Vehicles Exceeding Speed Limits by Ward
+#| echo: false
+
+# Summarize over speeding by ward
+ward_speeding_summary <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(
+    total_over_speeding = sum(over_speed_limit > 0, na.rm = TRUE),
+    avg_over_speeding = mean(over_speed_limit[over_speed_limit > 0], na.rm = TRUE)
+  )
+
+# Plot total over speeding by ward
+ggplot(ward_speeding_summary, aes(x = factor(ward_no), y = total_over_speeding, fill = total_over_speeding)) +
+  geom_col() +
+  scale_fill_gradient(low = "yellow", high = "red") +
+  labs(
+    x = "Ward Number",
+    y = "Total Over Speeding Incidents",
+    fill = "Over Speeding Volume"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+#| label: fig-cam-dens
+#| fig-cap: Camera Density vs. Speeding Behavior
+#| echo: false
+
+### Explore camera density per ward
+camera_density <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(cameras_nearby = n_distinct(X_id), .groups = "drop")
+
+ward_speed_analysis <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(
+    avg_speed_over_limit = mean(over_speed_limit, na.rm = TRUE),
+    cameras_nearby = n_distinct(X_id),
+    .groups = "drop"
+  )
+
+ggplot(ward_speed_analysis, aes(x = cameras_nearby, y = avg_speed_over_limit)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(
+       x = "Number of Cameras Nearby",
+       y = "Average Speed Over Limit")
+
+```
+
+```{r}
+#| label: fig-heatmap1
+#| fig-cap: Speeding Volume by Ward and Over Speed Limit
+#| echo: false
+
+# Aggregate data for heatmap
+heatmap_data <- final_filtered_data %>%
+  group_by(ward_no, over_speed_limit) %>%
+  summarize(total_volume = sum(volume, na.rm = TRUE)) %>%
+  ungroup()
+
+# Create the heatmap
+ggplot(heatmap_data, aes(x = factor(ward_no), y = over_speed_limit, fill = total_volume)) +
+  geom_tile(color = "white") +
+  scale_fill_gradient(low = "lightblue", high = "red", name = "Total Volume") +
+  labs(
+    x = "Ward Number",
+    y = "Over Speed Limit (km/h)"
+  ) +
+  theme_minimal() +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    panel.grid = element_blank()
+  )
+```
+
 # Model {#sec-mymodel}
+
+## Overview
+
+This section outlines the modeling process used to analyze over-speeding behavior based on data from Toronto’s School Safety Zone Watch Your Speed Program and Traffic Cameras datasets.
+The goal was to predict the extent to which observed speeds exceeded posted limits (over_speed_limit) by evaluating various predictors.
+Three models were developed and assessed using performance metrics, including RMSE, MAE, and \( R^2 \), to determine the most appropriate model.
+The final model selection was based on its balance between simplicity, accuracy, and interpretability.
+
+## Model Structure
+
+Common Variables Across All Models:
+
+- Speed Limit (x₁): Posted speed limit at a given location.
+
+- Volume (x₂): Number of vehicles observed in a speed bin.
+
+Model-Specific Variables:
+
+- Model 1: Includes ward-level categorical effects (ward_no) to account for geographic differences in speeding behavior.
+
+- Model 2: Omits ward-level effects to focus solely on speed limits and vehicle volumes.
+
+- Model 3: Adds a camera identifier (Xₓᵢd) to examine the influence of specific camera locations on speeding patterns.
+
+#### Model Formulations
+
+1. Model 1: Full Model with Ward-Level Effects
+   \[
+   y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \sum_{j=1}^{13} \beta_{3j} \text{ward\_no}_j + \epsilon
+   \]
+
+2. Model 2: Excluding Ward-Level Effects
+   \[
+   y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon
+   \]
+
+3. Model 3: Including Camera Identifiers
+   \[
+   y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \sum_{j=1}^{13} \beta_{3j} \text{ward\_no}_j + \beta_4 X_{id} + \epsilon
+   \]
+
+## Variable Selection and Final Model
+
+The selection process aimed to balance complexity and explanatory power.
+Ward-level effects were included to capture geographic variations, while Camera Identifiers were tested to evaluate their influence.
+Redundant variables were excluded to maintain simplicity.
+Model 1 was selected as the final model because it demonstrated superior performance across key metrics.
+By including ward-level effects, it effectively captured geographic differences in speeding behavior, making it the most suitable choice.
+
+## Model Validation and Diagnostics
+
+### Performance Metrics
+
+The RMSE for Model 1 was 7.01, indicating that, on average, predictions deviated from observed values by this amount.
+MAE was 4.77, highlighting its precision in predicting over-speeding behavior.
+The \( R^2 \) value of 0.176 showed that the model explained 17.6% of the variance in over-speeding behavior, providing a reasonable level of explanatory power for the observed data.
+
+### Residual Diagnostics
+
+Residual diagnostics confirmed the validity of Model 1.
+The residual standard error of 7.37 aligned closely with the RMSE, reinforcing the model’s accuracy.
+Residual plots exhibited no systematic patterns, supporting the assumption of linearity.
+Additionally, homoscedasticity was verified, indicating consistent variance in residuals across predictions.
+
+## Diagnostic Checks and Model Justification
+
+Variance Inflation Factors (VIFs) were calculated to evaluate multicollinearity, with all variables in Model 1 displaying acceptable values below 4.
+Although Model 3 included Camera Identifiers, this variable was statistically insignificant and did not enhance model performance.
+Model 1 effectively balanced complexity and accuracy, integrating ward-level effects to provide valuable insights into over-speeding behavior while maintaining interpretability.
+Consequently, Model 1 was selected as the final model.
 

### 2024-12-04 scripts/05-exploratory_analysis.R

@@ -82,6 +82,28 @@ ggplot(ward_speeding_summary, aes(x = factor(ward_no), y = total_speeding, fill 
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
                                      
 
+# Summarize over speeding by ward
+ward_speeding_summary <- final_filtered_data %>%
+  group_by(ward_no) %>%
+  summarize(
+    total_over_speeding = sum(over_speed_limit > 0, na.rm = TRUE),
+    avg_over_speeding = mean(over_speed_limit[over_speed_limit > 0], na.rm = TRUE)
+  )
+
+# Plot total over speeding by ward
+ggplot(ward_speeding_summary, aes(x = factor(ward_no), y = total_over_speeding, fill = total_over_speeding)) +
+  geom_col() +
+  scale_fill_gradient(low = "yellow", high = "red") +
+  labs(
+    title = "Number of Vehicles Exceeding Speed Limits by Ward",
+    x = "Ward Number",
+    y = "Total Over Speeding Incidents",
+    fill = "Over Speeding Volume"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+
 # Interactive map of speeding incidents
 leaflet(final_filtered_data %>% filter(speeding == 1)) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%

### 2024-12-04 paper/paper.qmd
final commit last checks and image trouble shooting. pls work

@@
 ---
-title: "Chasing Speed Demons: Analyzing Toronto’s Speeding Hotspots and Enforcement Gaps”"
-subtitle: "My subtitle if needed"
+title: "Chasing Speeders: Analyzing Toronto’s 2023 Speeding Trends in School Zones”"
+subtitle: "Volume Versus Speed Limits: 0.12 km/h Declines That Add Up Across School-Traffic Zones"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This study investigates over-speeding in Toronto, defined as the amount by which vehicles exceed posted speed limits, using data from the city's School Safety Zone Watch Your Speed Program and Traffic Camera datasets.
+A multi-linear regression analysis revealed key drivers of speeding behavior: higher traffic volumes reduce over-speeding by 0.12 km/h per additional vehicle, while increasing speed limits correlate with a 0.13 km/h rise in excess speed per km/h limit increase.
+Geographic variability was significant, with Ward 4 showing a 4.8 km/h reduction in over-speeding compared to the reference ward.
+Despite these findings, the negligible impact of speed cameras highlights limitations in existing enforcement data.
+These insights underscore the importance of nuanced, data-driven approaches to traffic management and policy development in urban settings."
 format: pdf
@@
 library(kableExtra)
+library(modelsummary)
+library(knitr)
+library(here)
 
 final_filtered_data <- read_parquet("../data/analysis_data/analysis_data.parquet")
+model1 <- readRDS("../models/first_model.rds")
 
@@
 
-This paper focuses on understanding the estimand **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit.
+This paper focuses on understanding the estimand **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit in 2023.
 This dependent variable is calculated as the difference between the lower bound of the observed speed bin and the speed limit, ensuring a straightforward measure of speeding severity.
@@
 ## Packages Used
-The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `kableExtra` [@kableExtra] and `rstanarm` [@rstanarm].
+The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `modelsummary` [@modelsummary], `kableExtra` [@kableExtra] and `rstanarm` [@rstanarm].
 
@@
 
-- **Over Speed Limit**: This key dependent variable quantifies the extent to which drivers exceed posted speed limits.
+- **Over Speed Limit**: This dependent variable quantifies the extent to which drivers exceed posted speed limits.
 It is derived by calculating the difference between the midpoint of the observed speed bin (e.g., `[45,50)` translates to a midpoint of 47.5 km/h) and the posted speed limit.
-This method provides a nuanced measure of speeding severity, distinguishing between minor infractions and extreme violations.
+This method provides a measure of speeding severity, distinguishing between minor infractions and extreme violations.
 
@@
 #| echo: false
+#| message: false
+#| warning: false
 
@@
 ```
+[fig-heatbar]  visualizes the total volume of over-speeding incidents across various wards.
+Ward 10 stands out prominently with the highest number of over-speeding incidents, exceeding 1,500, which may indicate unique characteristics such as higher traffic density or insufficient enforcement measures.
+Other wards, such as Wards 13 and 16, also show elevated levels of over-speeding but fall significantly behind Ward 10.
 
@@
 ```
+The scatter plot in [fig-cam-dens] explores the relationship between the number of speed cameras nearby and the average speed over the limit across various wards.
+The slight downward trend, indicated by the fitted regression line, suggests a weak negative association between the number of cameras and over-speeding behavior.
+Specifically, as the number of cameras increases, the average speed over the limit tends to decrease, albeit marginally.
+However, the variability in average speed over the limit across wards with the same number of cameras highlights the potential influence of other ward-specific factors, such as enforcement policies or road design.
 
@@
 
+[@fig-heatmap1] provides a detailed visualization of speeding behavior across Toronto’s wards, showcasing the relationship between the extent of over-speeding (measured in km/h above the limit) and the total volume of vehicles recorded.
+Ward 10 stands out with the highest total speeding volume, concentrated in lower over-speeding categories (<40 km/h above the limit), indicating significant non-compliance with posted limits.
+Conversely, instances of extreme over-speeding (>50 km/h above the limit) are sparse but occur across wards 4, 13, and 25, highlighting potential hotspots for dangerous driving behavior.
+
 # Model {#sec-mymodel}
@@
 
-- Speed Limit (x₁): Posted speed limit at a given location.
+- Speed Limit: Posted speed limit at a given location.
 
-- Volume (x₂): Number of vehicles observed in a speed bin.
+- Volume: Number of vehicles observed in a speed bin.
 
@@
 
-- Model 3: Adds a camera identifier (Xₓᵢd) to examine the influence of specific camera locations on speeding patterns.
+- Model 3: Adds a camera identifier (X_id) to examine the influence of specific camera locations on speeding patterns.
 
@@
 1. Model 1: Full Model with Ward-Level Effects
-   \[
-   y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \sum_{j=1}^{13} \beta_{3j} \text{ward\_no}_j + \epsilon
-   \]
+
+\begin{align}
+\text{over\_speed\_limit}_i &= \beta_0 + \beta_1 \times \text{speed\_limit}_i + \beta_2 \times \text{volume}_i + \sum_{j=1}^{13} \beta_{3j} \times \text{ward\_no}_{ij} + \epsilon_i
+\end{align}
 
 2. Model 2: Excluding Ward-Level Effects
-   \[
-   y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon
-   \]
 
-3. Model 3: Including Camera Identifiers
-   \[
-   y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \sum_{j=1}^{13} \beta_{3j} \text{ward\_no}_j + \beta_4 X_{id} + \epsilon
-   \]
+\begin{align}
+\text{over\_speed\_limit}_i &= \beta_0 + \beta_1 \times \text{speed\_limit}_i + \beta_2 \times \text{volume}_i + \epsilon_i
+\end{align}
+
+3. Model 3: Including Camera Identifiers (X_id)
+
+\begin{align}
+\text{over\_speed\_limit}_i &= \beta_0 + \beta_1 \times \text{speed\_limit}_i + \beta_2 \times \text{volume}_i + \sum_{j=1}^{13} \beta_{3j} \times \text{ward\_no}_{ij} + \beta_4 \times \text{X\_id}_i + \epsilon_i
+\end{align}
 
@@
 Redundant variables were excluded to maintain simplicity.
-Model 1 was selected as the final model because it demonstrated superior performance across key metrics.
-By including ward-level effects, it effectively captured geographic differences in speeding behavior, making it the most suitable choice.
+Model 1 was selected as the final model because it demonstrated superior performance across validation metrics.
+By including ward-level effects, it effectively captured geographic differences in speeding behaviour, making it the most suitable choice.
 
@@
 Additionally, homoscedasticity was verified, indicating consistent variance in residuals across predictions.
+Please see [@fig-residuals] in the appendix for a detailed graph.
 
@@
 
+## Model Summary Statistics
+The analysis examined over-speeding, defined as the amount (in km/h) by which vehicles exceeded the posted speed limit.
+In [@tbl-model-results], shows the variation in speeding effects for model 1, including the ward number, speed limit, and volume among other metrics.
+
+#### Analysis of Over-Speeding Behavior
+For every 1 km/h increase in the posted speed limit, the predicted over-speeding increased by 0.13 km/h, reflecting a slight inclination for drivers to exceed limits more on roads with higher speed thresholds.
+Traffic volume also played a role, with each additional vehicle in the observed volume leading to a decrease in over-speeding by 0.12 km/h.
+This suggests that heavier traffic may act as a natural deterrent to excessive speeding.
+
+Ward-level differences were also observed, with Ward 2 serving as the reference category.
+Ward 4, known for its high-speed enforcement on Parkside Drive, showed a reduction of 4.84 km/h in over-speeding compared to Ward 2.
+Larger reductions were observed in Ward 20 (10.84 km/h) and Ward 14 (11.22 km/h).
+Conversely, Ward 12 displayed an increase of 4.15 km/h, indicating that over-speeding was more prevalent in this area.
+
+#### Performance Statistics
+The model's residual standard error was 7.36 km/h, indicating the average deviation of predictions from observed speeding levels.
+This was supported by an RMSE of 7.36.
+A relative RMSE of 10.51% of the observed range and 32.91% of the standard deviation indicates moderate predictive accuracy.
+While it suggests the model captures a portion of the variability, the error may be considered substantial given the dataset's complexity.
+For urban traffic behavior, this level of error isn't ideal but isn't unexpected due to high variability and unobserved factors influencing over-speeding.
+
+The \(R^2\) value of 0.176 and adjusted \(R^2\) of 0.175 show that the model explains approximately 17.6% of the variance in over-speeding.
+Although relatively modest, this reflects the challenge of modeling complex urban traffic behaviors with diverse contributing factors.
+
+#### Statistical Fit Metrics
+With an AIC of 99,722.4 and a BIC of 99,851.4, the model, these values are consistent with the dataset's size, being 14,594 observations.The model's F-statistic of 207.459 shows the statistical significance of the predictors in explaining variations in over-speeding.
+This shows that the inclusion of speed limits, traffic volume, and ward-level effects contributes significantly to the model’s explanatory power.
+
+```{r}
+#| label: tbl-model-results
+#| tbl-cap: Summary of Regression Estimates for Model 1
+#| echo: false
+
+modelsummary(
+  list("Model 1" = model1),
+  fmt = 3,  # Number of decimals
+  gof_omit = ".*"  # Omit extra goodness-of-fit statistics if necessary
+)
+
+```
+
+## Relationship Between predicted and observed over-speeding values
+
+```{r}
+#| label: fig-observed-vs-predicted
+#| fig-cap: "Observed vs. Predicted Over-Speeding: Model Accuracy Visualization"
+
+knitr::include_graphics(here("other", "graphs", "predObsv.png"))
+
+```
+
+[@fig-observed-vs-predicted] shows the relationship between predicted and observed over-speeding values.
+Predicted values align reasonably well with observed values at lower ranges (0–10 km/h).
+This indicates that the model performs adequately in capturing moderate over-speeding behaviors, where most of the data points fall close to the red dashed line representing perfect predictions.
+
+### Underestimation of Higher Values
+For observed over-speeding values exceeding 20 km/h, the model consistently underestimates.
+Predicted values do not reach the observed extremes, suggesting that the model has difficulty accounting for the variability or extremity of high over-speeding incidents.
+This limitation is evident from the lack of points in the higher range of the dashed line.
+
+### Clustering at Lower Speeds
+There is a noticeable clustering of predictions around 0–10 km/h, even for observed over-speeding values as high as 40 km/h.
+This clustering suggests the model's predictions are conservative, failing to differentiate effectively between moderate and extreme over-speeding cases.
+
+While the model provides reasonable predictions for average over-speeding behaviors, its performance diminishes at higher observed values, indicating potential limitations in its ability to capture outliers or extreme behaviors.
+
+## Average Predicted Over-Speeding by Ward
+```{r}
+#| label: fig-ward-predictions
+#| fig-cap: Predicted Over-Speed Limit by Ward
+#| echo: false
+
+knitr::include_graphics(here("other", "graphs", "heatbarfigure.png"))
+
+```
+[@Fig-ward-predictions] shows the average predicted over-speeding across Toronto wards, measured in kilometers per hour.
+
+### Distribution of Predictions Across Wards
+Ward 12 exhibits the highest predicted over-speeding value, with an average of approximately 12.5 km/h over the speed limit.
+This value is notably higher compared to other wards, highlighting Ward 12 as a potential outlier.
+Similarly, Ward 25 follows closely with an average predicted over-speeding of about 11 km/h, suggesting consistently elevated predictions in these areas.
+
+### Wards with Lower Predictions
+In contrast, Wards 5, 21, and 20 show substantially lower predicted over-speeding values, averaging below 3 km/h.
+These results suggest that predicted over-speeding behaviors are less prominent in these wards, potentially influenced by traffic conditions, enforcement presence, or other unobserved factors.
+
+#### Implications of Predictions
+While the predictions align with observed patterns in some wards, such as higher over-speeding in Ward 12, the variability across wards suggests that spatial characteristics may play a significant role.
+These predictions provide a baseline understanding of over-speeding behavior at a ward level.
+
 # Discussion {#sec-mydisc}
 
-## First discussion point {#sec-first-point}
+## Ward-Level Disparities in Over-Speeding
+The results show geographical disparities in over-speeding across Toronto.
+For instance, Ward 12 emerges as a statistical outlier, with an average predicted over-speeding of 12.5 km/h.
+In  contrast, Ward 5 shows much lower predictions, averaging less than 3 km/h.
+This stark difference raises questions about factors such as road design, traffic enforcement, and even demographic variations.
+Ward 12 may be plagued by wide arterial roads inviting speed or does has outdated enforcement options.
+The low over-speeding in Ward 5 may hint at effective urban planning or stronger enforcement presence.
+It’s clear that “where” you drive in Toronto heavily influences “how” you drive.
+However, wards span cities in the 25 Ward model, and generalizing spaces of this size may also look over more specific area data.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Ward 4, which includes Parkside Drive—Toronto's most-ticketed location for speeding [@cbc2023], exhibited an intriguing pattern.
+Despite its reputation for dangerous speeding and expected dominance in over all speeding, the model indicated that over-speeding in Ward 4 was, on average, 4.77 km/h lower than the reference Ward 2.
+This discrepancy suggests the effectiveness of enforcement measures or specific traffic conditions in mitigating speeding behavior.
 
-## Second discussion point
+## Cameras
+While traffic cameras are often touted as a primary deterrent for speeding, their impact appears muted in the exploratory analysis.
+The inclusion of camera identifiers in regression model 3 did little to improve predictive accuracy, with the variable itself showing no statistical significance.
+Even the relationship between the number of cameras nearby and over-speeding is weak, challenging the assumption that simply adding cameras will curb speeding.
+Drivers may simply be accustomed to their presence or camera density may be too low to create a broader decreasing effect.
 
-## Third discussion point
+## Volume of Speeds
+Surprisingly, higher traffic volumes seem to correlate with lower over-speeding rates.
+The consistently negative coefficient for volume across all models suggests that when more vehicles are on the road at any speed, the likelihood of speeding diminishes.
+One possible explanation is that congestion creates natural constraints, limiting a driver’s ability to exceed the speed limit alongside the speed of their drivers around them.
 
-# Weaknesses and next steps {#sec-mylimits}
+## Outliers 
+The scatterplot of predicted vs.
+observed over-speeding values in [@fig-ward-predictions] reveals areas where the model struggles, particularly with extreme outliers.
+Predictions for high over-speeding values (e.g., above 40 km/h) are generally underrepresented, while predictions cluster heavily near lower values.
+This suggests the model may be oversimplifying the dynamics of extreme over-speeding events, which likely involve additional, unmeasured factors like driver behavior or road conditions.
+Additionally, the residual errors in [@fig-residuals] highlight slight over-predictions for certain wards, hinting at potential biases in how ward-level effects were estimated.
 
-Weaknesses and next steps should also be included.
+# Limitations and Next Steps {#sec-mylimits}
+
+## Weaknesses and Limitations
+
+### Data and Method Limitations
+The analysis does not account for date and time-specific patterns.
+Over-speeding frequently occurs during night hours or non-peak times when roads are less congested and school zone speeds only ooperate between 8 AM and 5 PM.
+Additionally, it is unclear whether the speed signs are specifically tracking school zone speeds or normal speeds.
+Some speed cameras included in the dataset were taken down partway through 2023, but they were retained in the analysis as long as they existed at any point during the year.
+While this decision did not significantly affect results, it did not clearly reflect temporal changes in camera presence.
+
+The School Safety Zone Watch Your Speed Program sensors may occasionally track outliers such as ambulances or police vehicles, leading to exaggerated over-speeding measurements.
+Additionally, these systems do not log vehicle types, which could have provided more support into driving patterns.
+The analysis assumes that vehicles maintain a constant speed through the sensor area, which may not reflect real-world driving behaviour.
+Traffic patterns can be complex and consequently difficult to accurately model.
+The multi-linear regression model used could benefit from more layered or bayesian models and uncertainty handling.
+
+A major assumption in the modeling process is the uniform distribution of speeds within speed bins.
+However, this may not reflect real-world conditions, where driver behavior often skews towards certain thresholds (e.g., staying just below enforcement limits).
+Similarly, this analysis aimed to examine trends in speeding around schools near traffic cameras, however it lacks in tracking speed demons and focuses on all speeds above the limit.
+
+### Geographic and Spatial Considerations
+
+Research suggests that proximity to schools and intersections significantly influences traffic flow.
+For instance, vehicle delays decrease when schools are located at least 400 meters from intersections [@mdpi2022].
+This analysis did not explicitly track specific spatial relationships and specific distances.
+Determining an appropriate spatial threshold for evaluating enforcement was solely based on the 400m reference.
+More research may be required to determine the adequacy of a 400-500m proximity.
+Furthermore, the School Safety Zone Watch Your Speed Program – Detailed Speed Counts dataset is over 12 months behind its refresh rate, limiting its applicability to current traffic conditions.
+
+## Next Steps and recommendations
+
+Future models could include weather data, as adverse conditions (e.g., rain, snow) are known to impact driving behavior.
+Open Data Toronto provides weather datasets that could be merged with speeding metrics for more comprehensive analyses, if joins are possible.
+Furthermore, considering the time of speed logging may improve and further detail the analysis, as speeding increases towards the peak hours and the night time [@nightSpeed] and it would better align with Toronto school zone speed laws from 8am-5pm.
+
+Although camera density was weakly correlated with over-speeding, as seen in the exploratory analysis, further research could evaluate enforcement alternatives, such as dynamic speed limits or enhanced police presence.
+This might also involve examining specific ticketing data to identify where enforcement is most effective.
+More detailed data including areas outside of school zones can offer a proper comparison to trends in current data exclusive to school zones.
+Camera-related variables were excluded from the final models due to multicollinearity.
+Addressing this with larger datasets or alternative feature engineering techniques may help retain these variables in future models while improving interpretability.
+
+Incorporating data on vehicle types could help identify whether certain vehicles (e.g., trucks, commercial vehicles) contribute disproportionately to over-speeding.
+Modified cars tend to speed at greater velocity and more often than others [@], offering additional insights into driver behaviour.
+Furthermore, there has been evidence that increases in vehicle height also correlate with speeding, as vehicle height may affect drivers' speed perception [@height].
+
+As speeding trends can vary greatly over time, addtional analysis spanning over several years can further support driver trends and behaviour.
+For example, speeding increased during the pandemic, likely due to the significant decrease in cars on the road, offering more leeway for speeding [@pandemic].
+Having a stronger focus on speed demons may be of interest, as tracking driver behaviour of those who minimally speed may not be as telling of general speeder habits.
 
@@
 
-If you're incorporating feedback to diversify question types and address previous gaps, here’s an improved version of your appendix for Toronto's speeding analysis with a reasonable budget.
-I'll also ensure it reflects insights from your earlier examples and feedback while incorporating poll aggregation and methodology details.
-
----
-
-# Idealized Survey & Methodology - $500K Budget {#sec-ideal-surv}
+# Idealized Survey & Methodology {#sec-ideal-surv}
 
 ## Overview
-This appendix outlines a survey methodology designed to enrich the analysis of excessive speeding near Toronto school zones.
-With a $500K budget, this approach aims to gather nuanced data from drivers, pedestrians, and community members about speeding behaviors, awareness of enforcement measures, and perceived safety.
-By integrating observational data with survey insights, this methodology enables robust analysis and informed policy recommendations.
-
----
+This appendix outlines a survey methodology designed to improve the analysis of excessive speeding near Toronto school zones.
+With a $500K budget, this approach aims to gather data from drivers, pedestrians, and community members about speeding behaviours, awareness of enforcement measures, and perceived safety.
+By joining observational data with survey insights, this methodology aims to improve analysis and informed policy recommendations.
 
 ## Budget Justification
-# Overview
 
-A $500,000 budget is proposed for analyzing speeding trends in Toronto school zones.
-Based on similar city-wide traffic safety initiatives, such as Vision Zero and speed enforcement studies, this budget is allocated to account for:
+Vision Zero’s annual budget for public engagement and data-driven initiatives often ranges between 1-2 million, justifying a scaled-down 500K budget for this specific study.Based on this similar city-wide traffic safety initiatives the $500K is allocated to account for:
 
-Comprehensive sampling: Targeting diverse demographics and road users.
+- Comprehensive sampling (targeting diverse demographics and road users).
 
-Advanced recruitment and incentives: Ensuring adequate participation.
+- Advanced recruitment and incentivization strategies to ensure adequate participation.
 
-High-quality data integration: Focused on cleaning, validation, and geospatial analysis.
-
-# Proportionality
-
-Vision Zero allocates $250,000 to $1,000,000 for focused safety measures annually, making $500,000 appropriate for a targeted study [@citytoronto2024].
-
-# Survey Costs
-
-Recruiting diverse participants typically costs $200,000 to $300,000, ensuring robust and representative data [@pew2023].
-
----
+- High-quality data cleaning, validation, and geospatial integration.
 
 ## Sampling Approach
-### **Target Groups**
-1. **Drivers:**  
-   - Key demographic: Ages 18-40, with emphasis on young male drivers.
+
+### Target Groups
+
+**Drivers:**  
+- Demographic: Ages 18-40, with emphasis on young male drivers.
+
 - Behavioral patterns: Frequent commuters in school zones.
 
-2. **Pedestrians:**  
+**Pedestrians:**  
 - Parents, school staff, and local residents.
 
-3. **Community Advocates:**  
+**Community Advocates:**  
 - Stakeholders involved in traffic safety initiatives.
 
-### **Sampling Method**
+### Sampling Method
+
 **Multi-Stage Sampling:**  
-1. **Stratification by Wards:**  
-   - Ensure geographic coverage across all Toronto wards with school zones.  
-2. **Cluster Sampling:**  
-   - Focus on areas near speed cameras and school safety zones.  
-3. **Random Sampling:**  
-   - Within clusters, randomly select participants for unbiased representation.
+- **Stratification by Wards:** Ensure geographic coverage across all Toronto wards with school zones.  
+- **Cluster Sampling:** Focus on areas near speed cameras and school safety zones.  
+- **Random Sampling:** Within clusters, randomly select participants for unbiased representation.
 
 **Sample Size Goal:**  
-- **3,000 respondents**: Sufficient to identify trends at the ward level while maintaining a margin of error below ±2%.
+3,000 respondents: Sufficient to identify trends at the ward level while maintaining a margin of error below ±2%.
 
----
+## Recruitment and Incentives
 
-## Recruitment Strategy
-### **Primary Channels**
-1. **Online Ads:**  
-   - Platforms: TikTok, Instagram, and YouTube.
-   - Geo-targeted to school zones and high-traffic areas.
-   - Budget Allocation: $150,000.
+### Recruitment Channels
+- **Digital Advertising:** Ads targeted toward young men, particularly those who frequent car meetups or automotive social media groups.
+Ads will use neutral language like "Toronto Traffic Safety Study" to avoid framing biases.
+  **Allocated Budget:** $200,000.  
+- **Community Partnerships:** Collaborations with schools, PTAs, and driving schools to involve families and new drivers.  
+  **Allocated Budget:** $50,000.
 
-2. **On-the-Ground Recruitment:**  
-   - Flyers distributed in school zones, community centers, and traffic courts.
-   - Budget Allocation: $50,000.
+### Incentives
+- A raffle for ten $50 gift cards to encourage participation.  
+- Highlighting confidentiality ensures respondents feel safe sharing honest information.
 
-3. **Partnerships with Apps:**  
-   - Collaboration with navigation tools like Waze to deliver in-app survey prompts.
-   - Budget Allocation: $100,000.
+### Behavioral Targeting
+Beyond demographic data, the recruitment strategy will aim to identify drivers who:
 
-4. **Incentives:**  
-   - Monetary rewards: $20 gift cards for completion, with a chance to win larger prizes.
-   - Budget Allocation: $100,000.
+- Drive frequently during peak hours near school zones.
 
----
+- Own or lease modified vehicles.
 
-## Survey Design
+- Exhibit risky behaviors, such as speeding or tailgating, based on self-reporting and survey insights.
 
-### Diversified Question Types
-To address the feedback, the survey incorporates **varied question formats**:
+- Young males (18–34), a group overrepresented in speeding violations [@transportcanada2019].
 
-1. **Multiple Choice:**  
-   - Example: "What factors most influence your speed in school zones?"
-     - Time pressure.
-     - Peer behavior.
-     - Lack of awareness of speed limits.
-     - Others.
+- Drivers who own modified cars, as these factors are linked to increased risk of speeding [@safetyresearch2018].
 
-2. **Likert Scales:**  
-   - Example: "Rate your agreement with: 'Speed cameras make me feel safer in school zones.'"
-     - Strongly disagree to strongly agree.
+## Bias Mitigation
 
-3. **Open-Ended:**  
-   - Example: "What other measures would help reduce speeding in school zones?"
-
-4. **Ranking Questions:**  
-   - Example: "Rank the following in order of their effectiveness at reducing speeding: speed cameras, visible signage, increased fines, police presence."
-
-5. **Interactive Scenario Questions:**  
-   - Example: "If you saw a speed camera sign but no visible camera, how likely are you to still speed? (Very unlikely to very likely)."
-
----
-
-### Survey Flow
-1. **Introduction and Consent:**  
-   - Explain the purpose, time required (5-7 minutes), and confidentiality.
-
-2. **Demographics:**  
-   - Age, gender, driving habits, and proximity to school zones.
-
-3. **Behavioral Insights:**  
-   - Driving patterns, awareness of speed limits, and interaction with enforcement measures.
-
-4. **Perceived Effectiveness:**  
-   - Opinions on speed cameras, signage, and enforcement presence.
-
-5. **Safety Concerns:**  
-   - Perceptions of safety in school zones and suggested improvements.
-
----
+1. **Survey Naming:** The neutral title **Toronto Traffic Safety Study** reduces response bias by framing the study as a general exploration of driver behavior rather than focusing on speeding.
+2. **Confidentiality Assurance:** Respondents are explicitly informed that:
+   - Data is anonymous and used solely for research purposes.
+   - Identifiable information (e.g., email for raffle entries) is stored separately.
+3. **Question Design:** Questions are carefully ordered to build trust, with sensitive topics (e.g., fastest speed driven) placed later in the survey.
+4. **Response Validation:** Attention-check questions ensure participant engagement and accuracy.
 
@@
   - **Proximity to enforcement measures.**
+
   - **Frequency of driving in school zones.**
+
 - Integrate responses with speed count data to build predictive models of speeding behavior.
-
-### **Modeling Framework**
-- **Dependent Variable:** Likelihood of exceeding speed limits.  
-- **Predictors:**  
-  - Awareness of enforcement measures.  
-  - Demographics (age, gender).  
-  - Self-reported speeding behavior.  
-  - Proximity to cameras and speed signs.
-
-### **Diagnostics**
-- Residual analysis and goodness-of-fit tests to validate model performance.
-- Q-Q plots and distribution checks for robustness.
-
----
-
-## Budget Breakdown
-
-| **Expense**                     | **Cost Estimate (CAD)**  |
-|----------------------------------|--------------------------|
-| Social Media Recruitment        | $150,000                 |
-| In-App Ads and Recruitment      | $100,000                 |
-| On-the-Ground Recruitment       | $50,000                  |
-| Participant Incentives          | $100,000                 |
-| Data Cleaning and Validation    | $75,000                  |
-| Survey Design and Implementation| $25,000                  |
-| **Total**                       | **$500,000**             |
 
@@
 1. **Survey Link:**  
-   [Insert Google Forms Link Here].
+   [https://forms.gle/TbRYwoGafy2EzwBQ9].
 
-2. **Copy of Survey Questions:**  
-   - Include all questions, organized by section.
+2. **Full Survey Questions and Overview:**
 
-3. **Model Diagnostics:**  
-   - Residual plots, fit comparisons, and distribution checks.
+**Title:** Toronto Traffic Safety Study  
+**Purpose:** To study driver behavior near school zones.  
+**Estimated Time:** Less than 5 minutes.  
+**Confidentiality:** Responses are anonymous and used solely for research.
+
+#### Section 1: Demographics and Background
+1. What is your age group?  
+   - Under 18  
+   - 18–24  
+   - 25–34  
+   - 35–44  
+   - 45–54  
+   - 55–64  
+   - 65+
+
+2. What is your gender identity?  
+   - Male  
+   - Female  
+   - Non-binary  
+   - Prefer not to say
+
+3. What is your primary occupation?  
+   - Full-time student  
+   - Part-time student  
+   - Employed full-time  
+   - Employed part-time  
+   - Unemployed  
+   - Retired
+
+4. Do you currently own or lease a vehicle?  
+   - Yes, I own  
+   - Yes, I lease  
+   - No, I do not own or lease a car
+
+5. How old is your primary vehicle?  
+   - Less than 1 year  
+   - 1–3 years  
+   - 4–6 years  
+   - 7–10 years  
+   - Over 10 years
+
+#### Section 2: Driving Habits
+6. How often do you drive near schools in Toronto?  
+   - Daily  
+   - A few times a week  
+   - Rarely  
+   - Never
+
+7. What time of day do you most often drive near school zones?  
+   - Morning (6 AM – 9 AM)  
+   - Afternoon (12 PM – 3 PM)  
+   - Evening (4 PM – 7 PM)  
+   - Late night (after 9 PM)
+
+8. Do you typically drive alone or with others?  
+   - Alone  
+   - With friends or family
+
+#### Section 3: Speeding and Safety
+9. What is the fastest you’ve ever driven (in km/h)? *(Select closest range)*  
+   - Below 80 km/h  
+   - 80–99 km/h  
+   - 100–120 km/h  
+   - Over 120 km/h
+
+10. What do you consider a normal amount to speed over the limit (in km/h)? *(Select closest range)*  
+    - 0–5 km/h  
+    - 6–10 km/h  
+    - 11–20 km/h  
+    - 21+ km/h
+
+11. Do you think there are valid reasons to speed? *(Open-ended)*
+
+12. How likely are you to slow down when you see a speed camera or radar sign?  
+    - Always  
+    - Often  
+    - Sometimes  
+    - Rarely  
+    - Never
+
+#### Section 4: Traffic Safety Knowledge
+13. **Test Question:** What is the legal speed limit near school zones in Toronto?  
+    - 30 km/h  
+    - 40 km/h  
+    - 50 km/h  
+    - I don’t know
+
+#### Section 5: Traffic Incidents
+14. Have you ever been involved in a collision while driving?  
+    - Yes  
+    - No
+
+15. If yes, what was the estimated damage?  
+    - Below $500  
+    - $500–$1,000  
+    - Over $1,000
+
+#### Section 6: Raffle Entry
+16. Would you like to enter the raffle for a $50 gift card? *(Optional, email required)*  
+    - Yes (Enter email address)  
+    - No
 
 ---
+# Residual Plot for model 1
 
-This improved version addresses the feedback, diversifies question types, and integrates poll aggregation and modeling. Let me know if you’d like refinements!
+```{r}
+#| label: fig-residuals
+#| fig-cap: Residuals vs. Fitted Values for Model 1
+#| echo: false
 
+knitr::include_graphics(here("other", "graphs", "residuals.png"))
+
+```
 \newpage

### 2024-12-07 paper/paper.qmd
improve/fix wording, formatting, extra text

@@
 abstract: "This study investigates over-speeding in Toronto, defined as the amount by which vehicles exceed posted speed limits, using data from the city's School Safety Zone Watch Your Speed Program and Traffic Camera datasets.
-A multi-linear regression analysis revealed key drivers of speeding behavior: higher traffic volumes reduce over-speeding by 0.12 km/h per additional vehicle, while increasing speed limits correlate with a 0.13 km/h rise in excess speed per km/h limit increase.
+A multi-linear regression analysis revealed drivers of speeding behaviour near schools: higher traffic volumes reduce over-speeding by 0.12 km/h per additional vehicle, while increasing speed limits correlate with a 0.13 km/h rise in excess speed per km/h limit increase.
 Geographic variability was significant, with Ward 4 showing a 4.8 km/h reduction in over-speeding compared to the reference ward.
 Despite these findings, the negligible impact of speed cameras highlights limitations in existing enforcement data.
-These insights underscore the importance of nuanced, data-driven approaches to traffic management and policy development in urban settings."
+It highlights the need for better data to address traffic challenges effectively.
+While the results offer some direction, more comprehensive datasets could lead to stronger and more actionable policy recommendations."
 format: pdf
@@
 The rise of advanced vehicle technology and the adrenaline appeal of street racing exacerbate these issues, with young male drivers (18–25 years) disproportionately represented in this trend.
-Though rare at **0.9% prevalence**, street racing poses a **five-fold increase in crash risk (OR = 5.23, p < .001)**, even after accounting for other risk factors [@simonsmorton2017].
-Despite Ontario’s strict penalties, such as **$10,000 fines** and **30-day license suspensions for excessive speeding**, enforcement alone has not fully addressed these challenges [@trafficparalegal2024].
-Recent headlines, such as a driver clocked at **102 km/h in a 50 km/h school zone** and a Toronto speed camera being destroyed less than 24 hours after reinstallation, underscore the disconnect between enforcement measures and public behavior [@citynews2024; @cbc2023].
+Though rare at 0.9% prevalence, street racing poses a five-fold increase in crash risk (OR = 5.23, p < .001), even after accounting for other risk factors [@simonsmorton2017].
+Despite Ontario’s strict penalties, such as $10,000 fines and 30-day license suspensions for excessive speeding, enforcement alone has not fully addressed these challenges [@trafficparalegal2024].
+Recent headlines, such as a driver clocked at **102 km/h in a 50 km/h school zone and a Toronto speed camera being destroyed less than 24 hours after reinstallation, showing the disconnect between enforcement measures and public behaviour [@citynews2024; @cbc2023].
 
-This paper focuses on understanding the estimand **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit in 2023.
+This paper focuses on understanding the estimand over_speed_limit, the amount by which speeds in each observed speed bin exceed the posted speed limit around schools in 2023.
 This dependent variable is calculated as the difference between the lower bound of the observed speed bin and the speed limit, ensuring a straightforward measure of speeding severity.
@@
 To address this, the study employs multi-linear regression models to examine the relationship between excess speed and predictors such as traffic volume, speed limits, and ward characteristics throughout 2023.
-Results indicate that higher traffic volume reduces excess speed by **0.12 km/h per additional vehicle**, while speed limits positively influence speeding by **0.13 km/h per unit increase**.
+Results indicate that higher traffic volume reduces excess speed by 0.12 km/h per additional vehicle, while speed limits positively influence speeding by 0.13 km/h per unit increase.
 Wards exhibit substantial variability, with some, such as Ward 4, showing significant decreases in speeding.
@@
 The data for this analysis is sourced from Open Data Toronto [@opendatatoronto] and comprises three datasets: School Safety Zone Watch Your Speed Program – Locations [@schoolzonelocations], School Safety Zone Watch Your Speed Program – Detailed Speed Counts in 2023 [@schoolzonespeeds], and the Traffic Camera dataset  [@trafficcameras].
-These datasets provide detailed insights into speeding behavior, traffic enforcement mechanisms, and geographic characteristics within Toronto.
-The traffic camera and speed sign data were merged through a spacial join with a range of 500m using geographical coordinates.
+These datasets provide direction into speeding behaviour near schools, traffic enforcement mechanisms, and geographic characteristics within Toronto.
+The traffic camera and speed sign data were merged through a spatial join with a range of 500m using geographical coordinates.
 This was joined with the detailed counts of speeds through Toronto ward number (25 ward structure).
@@
 ## Packages Used
-The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `modelsummary` [@modelsummary], `kableExtra` [@kableExtra] and `rstanarm` [@rstanarm].
+The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `modelsummary` [@modelsummary], `kableExtra` [@kableExtra], `knitr` [@knitr], `here` [@here], and `rstanarm` [@rstanarm].
 
@@
 
-The datasets from Open Data Toronto serve as the foundation for analyzing speeding behavior in Toronto.
-Each variable is derived from real-world phenomena and represents a specific aspect of traffic patterns or enforcement mechanisms:
-
-- **Over Speed Limit**: This dependent variable quantifies the extent to which drivers exceed posted speed limits.
+- **over_speed_limit**: This dependent variable quantifies the extent to which drivers exceed posted speed limits.
 It is derived by calculating the difference between the midpoint of the observed speed bin (e.g., `[45,50)` translates to a midpoint of 47.5 km/h) and the posted speed limit.
@@
 
-- **Sign ID and Camera ID (X_id)**: These unique identifiers represent speed cameras and their locations, facilitating the mapping of speeding patterns to enforcement points.
-However, sign addresses may not always align precisely with camera placements, as discussed in the limitations section.
+- **sign_id and X_id (Camera id)**: These unique identifiers represent speed cameras and their locations near schools, However, sign addresses may not always align precisely with camera placements, as discussed in the limitations section.
 
-- **Longitude and Latitude**: Geographic coordinates capturing the locations of observed speeding behaviour, allowing for spatial analysis and the identification of high-risk zones.
+- **longitude and latitude**: Geographic coordinates capturing the locations of observed speeding behaviour, allowing for spatial analysis and the identification of high-risk zones.
 
-- **Speed Limit**: The posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h. This variable provides the baseline against which speeding is measured.
+- **speed_limit**: The posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h. This variable provides the baseline against which speeding is measured.
 
-- **Volume**: The count of vehicles observed in a given speed bin, representing traffic density. Median volume is 13 vehicles, with counts ranging widely, up to a maximum of 209 vehicles.
+- **volume**: The count of vehicles observed in a given speed bin, representing traffic density. The median volume is 13 vehicles, with counts ranging widely, up to a maximum of 209 vehicles.
 
-- **Speed Bin**: Recorded ranges of observed speeds, such as `[45,50)`, provide context for speed categories, assuming uniform speed distributions within each bin.
+- **speed_bin**: Recorded ranges of observed speeds, such as `[45,50)`, provide context for speed categories, assuming uniform speed distributions within each bin.
 
-- **Ward Number**: Geographic zones represented as categorical identifiers, reflecting administrative boundaries within Toronto.
+- **ward_no**: Geographic wards represented as 25 categorical identifiers, reflecting administrative boundaries within Toronto.
 For instance, Ward 10 accounts for 28.8% of observations, making it the most represented ward, while Ward 19 has considerably fewer observations.
 
-- **No Camera in Radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations.
+- **no_camera_in_radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations.
 If a given location fell outside the defined radius of nearby cameras, it was flagged as `TRUE`.
-This captures the absence of enforcement presence and allows for an analysis of its influence on speeding behavior.
+This captures the absence of enforcement presence and allows for an analysis of its influence on speeding behaviour.
+This was exclusively used during simulation.
 
-- **Camera Density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward.
-This variable offers insight into how concentrated enforcement might correlate with traffic patterns or speeding violations.
+- **camera_density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward.
+This variable attempts to show how concentrated enforcement might correlate with traffic patterns or speeding violations.
 
@@
 The datasets used in this study have notable limitations that must be acknowledged.
-**Sign addresses** in the datasets have not been verified and may represent the address of an adjacent property rather than the actual location of the sign.
-The **speed limit** column reflects the most recent operating parameter, which does not necessarily align with the current or previous speed limits for each location.
+Sign addresses in the datasets have not been verified and may represent the address of an adjacent property rather than the actual location of the sign.
+The speed limit column reflects the most recent operating parameter, which does not necessarily align with the current or previous speed limits for each location.
 
-The **Traffic Cameras** dataset is highly up-to-date, with a last refresh on 2024-11-26.
+The Traffic Cameras dataset is highly up-to-date, with a last refresh on 2024-11-26.
 While there are no issues with accessibility or completeness, usability is constrained by unclear column names and incomplete metadata.
-Conversely, the **School Safety Zone Watch Your Speed Program – Detailed Speed Counts** dataset is over 12 months outdated, requiring reliance on direct CSV links for access.
-This reduces its applicability for analysing current traffic trends.
+Conversely, the School Safety Zone Watch Your Speed Program – Detailed Speed Counts dataset is over 12 months outdated, requiring reliance on direct CSV links for access.
+This reduces its applicability for analyzing current traffic trends.
 
-To address gaps, **spatial joins** were employed to map camera locations to speed observation points, identifying enforcement presence, and **calculated variables**, such as over speed limit and camera density, were derived to facilitate analysis.
+To address gaps, spatial joins were employed to map camera locations to speed observation points, identifying enforcement presence, and calculated variables, such as over speed limit and camera density, were derived to facilitate analysis.
 However, these transformations have limitations.
@@
 ## Descriptive Statistics
-Descriptive statistics in [@tbl-summary] highlight critical patterns within the data:
-
-- The **mean over speed limit** is 3.91 km/h, with extreme violations reaching up to 70 km/h.
-
-- The **median speed bin midpoint** is 30 km/h, while the mean traffic volume is 23.35 vehicles.
-
-- Geographic variables such as **longitude** and **latitude** center around Toronto’s urban core.
+The descriptive statistics (see [@tbl-summary]) reveal important patterns within the dataset.
+The average over-speeding value is 3.91 km/h, with extreme cases reaching as high as 70 km/h, illustrating the range of speeding behavior observed.
+The median speed bin midpoint is 30 km/h, reflecting typical speed categories captured, while the mean traffic volume stands at 23.35 vehicles, indicating the density of traffic during observations.
+Geographic variables, including longitude and latitude, align closely with Toronto's urban core, emphasizing the city-centric focus of the analysis.
 
@@
 ```
-[fig-heatbar]  visualizes the total volume of over-speeding incidents across various wards.
+[@fig-heatbar]  visualizes the total volume of over-speeding incidents across various wards.
 Ward 10 stands out prominently with the highest number of over-speeding incidents, exceeding 1,500, which may indicate unique characteristics such as higher traffic density or insufficient enforcement measures.
@@
 ```
-The scatter plot in [fig-cam-dens] explores the relationship between the number of speed cameras nearby and the average speed over the limit across various wards.
-The slight downward trend, indicated by the fitted regression line, suggests a weak negative association between the number of cameras and over-speeding behavior.
+The scatter plot in [@fig-cam-dens] explores the relationship between the number of speed cameras nearby and the average speed over the limit across various wards.
+The slight downward trend, indicated by the fitted regression line, suggests a weak negative association between the number of cameras and over-speeding behaviour.
 Specifically, as the number of cameras increases, the average speed over the limit tends to decrease, albeit marginally.
@@
 
-[@fig-heatmap1] provides a detailed visualization of speeding behavior across Toronto’s wards, showcasing the relationship between the extent of over-speeding (measured in km/h above the limit) and the total volume of vehicles recorded.
+[@fig-heatmap1] provides a detailed visualization of speeding behaviour across Toronto’s wards, showcasing the relationship between the extent of over-speeding (measured in km/h above the limit) and the total volume of vehicles recorded.
 Ward 10 stands out with the highest total speeding volume, concentrated in lower over-speeding categories (<40 km/h above the limit), indicating significant non-compliance with posted limits.
-Conversely, instances of extreme over-speeding (>50 km/h above the limit) are sparse but occur across wards 4, 13, and 25, highlighting potential hotspots for dangerous driving behavior.
+Conversely, instances of extreme over-speeding (>50 km/h above the limit) are sparse but occur across wards 4, 13, and 25, highlighting potential hotspots for dangerous driving behaviour.
 
@@
 
-This section outlines the modeling process used to analyze over-speeding behavior based on data from Toronto’s School Safety Zone Watch Your Speed Program and Traffic Cameras datasets.
+This section outlines the modelling process used to analyze over-speeding behaviour based on data from Toronto’s School Safety Zone Watch Your Speed Program and Traffic Cameras datasets.
 The goal was to predict the extent to which observed speeds exceeded posted limits (over_speed_limit) by evaluating various predictors.
@@
 
-- Model 1: Includes ward-level categorical effects (ward_no) to account for geographic differences in speeding behavior.
+- Model 1: Includes ward-level categorical effects (ward_no) to account for geographic differences in speeding behaviour.
 
@@
 The selection process aimed to balance complexity and explanatory power.
-Ward-level effects were included to capture geographic variations, while Camera Identifiers were tested to evaluate their influence.
+Ward-level effects were included to capture geographic variations, while camera Identifiers were tested to evaluate their influence.
 Redundant variables were excluded to maintain simplicity.
@@
 The RMSE for Model 1 was 7.01, indicating that, on average, predictions deviated from observed values by this amount.
-MAE was 4.77, highlighting its precision in predicting over-speeding behavior.
-The \( R^2 \) value of 0.176 showed that the model explained 17.6% of the variance in over-speeding behavior, providing a reasonable level of explanatory power for the observed data.
+MAE was 4.77, highlighting its precision in predicting over-speeding behaviour.
+The \( R^2 \) value of 0.176 showed that the model explained 17.6% of the variance in over-speeding behaviour, providing a reasonable level of explanatory power for the observed data.
 
@@
 Variance Inflation Factors (VIFs) were calculated to evaluate multicollinearity, with all variables in Model 1 displaying acceptable values below 4.
-Although Model 3 included Camera Identifiers, this variable was statistically insignificant and did not enhance model performance.
-Model 1 effectively balanced complexity and accuracy, integrating ward-level effects to provide valuable insights into over-speeding behavior while maintaining interpretability.
+Although Model 3 included camera Identifiers, this variable was statistically insignificant and did not enhance model performance.
+Model 1 effectively balanced complexity and accuracy, integrating ward-level effects to show over-speeding behaviour while maintaining interpretability.
 Consequently, Model 1 was selected as the final model.
@@
 
-The \(R^2\) value of 0.176 and adjusted \(R^2\) of 0.175 show that the model explains approximately 17.6% of the variance in over-speeding.
-Although relatively modest, this reflects the challenge of modeling complex urban traffic behaviors with diverse contributing factors.
+The \(R^2\) value of 0.176 and adjusted \(R^2\) of 0.175 shows that the model explains approximately 17.6% of the variance in over-speeding.
+Although relatively modest, this reflects the challenge of modelling complex urban traffic behaviours with diverse contributing factors.
 
 #### Statistical Fit Metrics
-With an AIC of 99,722.4 and a BIC of 99,851.4, the model, these values are consistent with the dataset's size, being 14,594 observations.The model's F-statistic of 207.459 shows the statistical significance of the predictors in explaining variations in over-speeding.
+With an AIC of 99,722.4 and a BIC of 99,851.4, the model, values are consistent with the dataset's size, being 14,594 observations.
+The model's F-statistic of 207.459 shows the statistical significance of the predictors in explaining variations in over-speeding.
 This shows that the inclusion of speed limits, traffic volume, and ward-level effects contributes significantly to the model’s explanatory power.
@@
 #| fig-cap: "Observed vs. Predicted Over-Speeding: Model Accuracy Visualization"
+#| echo: false
 
@@
 Predicted values align reasonably well with observed values at lower ranges (0–10 km/h).
-This indicates that the model performs adequately in capturing moderate over-speeding behaviors, where most of the data points fall close to the red dashed line representing perfect predictions.
+This indicates that the model performs adequately in capturing moderate over-speeding behaviours, where most of the data points fall close to the red dashed line representing perfect predictions.
 
@@
 
-While the model provides reasonable predictions for average over-speeding behaviors, its performance diminishes at higher observed values, indicating potential limitations in its ability to capture outliers or extreme behaviors.
+While the model provides reasonable predictions for average over-speeding behaviours, its performance diminishes at higher observed values, indicating potential limitations in its ability to capture outliers or extreme behaviours.
 
@@
 In contrast, Wards 5, 21, and 20 show substantially lower predicted over-speeding values, averaging below 3 km/h.
-These results suggest that predicted over-speeding behaviors are less prominent in these wards, potentially influenced by traffic conditions, enforcement presence, or other unobserved factors.
+These results suggest that predicted over-speeding behaviours are less prominent in these wards, potentially influenced by traffic conditions, enforcement presence, or other unobserved factors.
 
@@
 While the predictions align with observed patterns in some wards, such as higher over-speeding in Ward 12, the variability across wards suggests that spatial characteristics may play a significant role.
-These predictions provide a baseline understanding of over-speeding behavior at a ward level.
+These predictions provide a baseline understanding of over-speeding behaviour at a ward level.
 
@@
 ## Ward-Level Disparities in Over-Speeding
-The results show geographical disparities in over-speeding across Toronto.
+The results show geographical disparities in over-speeding across Toronto school areas.
 For instance, Ward 12 emerges as a statistical outlier, with an average predicted over-speeding of 12.5 km/h.
@@
 This stark difference raises questions about factors such as road design, traffic enforcement, and even demographic variations.
-Ward 12 may be plagued by wide arterial roads inviting speed or does has outdated enforcement options.
+Ward 12 may be plagued by wide arterial roads inviting speed or does have outdated enforcement options.
 The low over-speeding in Ward 5 may hint at effective urban planning or stronger enforcement presence.
@@
 Ward 4, which includes Parkside Drive—Toronto's most-ticketed location for speeding [@cbc2023], exhibited an intriguing pattern.
-Despite its reputation for dangerous speeding and expected dominance in over all speeding, the model indicated that over-speeding in Ward 4 was, on average, 4.77 km/h lower than the reference Ward 2.
-This discrepancy suggests the effectiveness of enforcement measures or specific traffic conditions in mitigating speeding behavior.
+Despite its reputation for dangerous speeding and expected dominance in overall speeding, the model indicated that over-speeding in Ward 4 was, on average, 4.77 km/h lower than the reference Ward 2.
+This discrepancy suggests the effectiveness of enforcement measures or specific traffic conditions in mitigating speeding behaviour.
 
@@
 ## Volume of Speeds
-Surprisingly, higher traffic volumes seem to correlate with lower over-speeding rates.
+Higher traffic volumes seem to correlate with lower over-speeding rates.
 The consistently negative coefficient for volume across all models suggests that when more vehicles are on the road at any speed, the likelihood of speeding diminishes.
-One possible explanation is that congestion creates natural constraints, limiting a driver’s ability to exceed the speed limit alongside the speed of their drivers around them.
+One possible explanation is that congestion creates natural constraints, limiting a driver’s ability to exceed the speed limit alongside the speed of the drivers around them.
 
@@
 Predictions for high over-speeding values (e.g., above 40 km/h) are generally underrepresented, while predictions cluster heavily near lower values.
-This suggests the model may be oversimplifying the dynamics of extreme over-speeding events, which likely involve additional, unmeasured factors like driver behavior or road conditions.
+This suggests the model may be oversimplifying the dynamics of extreme over-speeding events, which likely involve additional, unmeasured factors like driver behaviour or road conditions.
 Additionally, the residual errors in [@fig-residuals] highlight slight over-predictions for certain wards, hinting at potential biases in how ward-level effects were estimated.
@@
 The analysis does not account for date and time-specific patterns.
-Over-speeding frequently occurs during night hours or non-peak times when roads are less congested and school zone speeds only ooperate between 8 AM and 5 PM.
+Over-speeding frequently occurs during night hours or non-peak times when roads are less congested and school zone speeds only operate between 8 AM and 5 PM.
 Additionally, it is unclear whether the speed signs are specifically tracking school zone speeds or normal speeds.
@@
 Traffic patterns can be complex and consequently difficult to accurately model.
-The multi-linear regression model used could benefit from more layered or bayesian models and uncertainty handling.
+The multi-linear regression model used could benefit from more layered or Bayesian models and uncertainty handling.
 
-A major assumption in the modeling process is the uniform distribution of speeds within speed bins.
-However, this may not reflect real-world conditions, where driver behavior often skews towards certain thresholds (e.g., staying just below enforcement limits).
+A major assumption in the modelling process is the uniform distribution of speeds within speed bins.
+However, this may not reflect real-world conditions, where driver behaviour often skews towards certain thresholds (e.g., staying just below enforcement limits).
 Similarly, this analysis aimed to examine trends in speeding around schools near traffic cameras, however it lacks in tracking speed demons and focuses on all speeds above the limit.
@@
 
-## Next Steps and recommendations
+## Next Steps and Recommendations
 
-Future models could include weather data, as adverse conditions (e.g., rain, snow) are known to impact driving behavior.
-Open Data Toronto provides weather datasets that could be merged with speeding metrics for more comprehensive analyses, if joins are possible.
-Furthermore, considering the time of speed logging may improve and further detail the analysis, as speeding increases towards the peak hours and the night time [@nightSpeed] and it would better align with Toronto school zone speed laws from 8am-5pm.
+Future models could include weather data, as adverse conditions (e.g., rain, snow) are known to impact driving behaviour.
+Open Data Toronto provides weather datasets that could be merged with speeding metrics for more comprehensive analyses if joins are possible.
+Furthermore, wards with greater populations could have weight added to them according to size.
 
@@
 Incorporating data on vehicle types could help identify whether certain vehicles (e.g., trucks, commercial vehicles) contribute disproportionately to over-speeding.
-Modified cars tend to speed at greater velocity and more often than others [@], offering additional insights into driver behaviour.
+Modified cars may tend to speed at greater velocity and more often than others due to adjusted, increased speed capacity, offering additional information on driver behaviour.
 Furthermore, there has been evidence that increases in vehicle height also correlate with speeding, as vehicle height may affect drivers' speed perception [@height].
 
-As speeding trends can vary greatly over time, addtional analysis spanning over several years can further support driver trends and behaviour.
+As speeding trends can vary greatly over time, additional analysis spanning over several years can further support driver trends and behaviour.
 For example, speeding increased during the pandemic, likely due to the significant decrease in cars on the road, offering more leeway for speeding [@pandemic].
 Having a stronger focus on speed demons may be of interest, as tracking driver behaviour of those who minimally speed may not be as telling of general speeder habits.
+Furthermore, considering the time of speed logging may improve and further detail the analysis, as speeding increases towards the peak hours and the night time [@nightSpeed] and it would better align with Toronto school zone speed laws from 8 am-5 pm.
 
@@
 With a $500K budget, this approach aims to gather data from drivers, pedestrians, and community members about speeding behaviours, awareness of enforcement measures, and perceived safety.
-By joining observational data with survey insights, this methodology aims to improve analysis and informed policy recommendations.
+By joining observational data with survey data, this methodology aims to improve analysis and informed policy recommendations.
 
@@
 
-Vision Zero’s annual budget for public engagement and data-driven initiatives often ranges between 1-2 million, justifying a scaled-down 500K budget for this specific study.Based on this similar city-wide traffic safety initiatives the $500K is allocated to account for:
+Vision Zero’s annual budget for public engagement and data-driven initiatives often ranges between 1-2 million, justifying a scaled-down 500K budget for this specific study.
+Based on these similar city-wide traffic safety initiatives the $500K is allocated to account for the following:
 
@@
 
-- Advanced recruitment and incentivization strategies to ensure adequate participation.
+- Thorough recruitment and incentivization strategies to ensure adequate participation.
 
@@
 
-**Drivers:**  
-- Demographic: Ages 18-40, with emphasis on young male drivers.
+Drivers, particularly those aged 18-40 with an emphasis on young male drivers, are included to understand the behavior of frequent commuters in school zones.
+Pedestrians, such as parents, school staff, and local residents, provide analytics into the community's experiences and concerns.
+Finally, community advocates, including stakeholders involved in traffic safety initiatives, contribute broader perspectives on strategies to improve safety in high-risk areas.
 
-- Behavioral patterns: Frequent commuters in school zones.
+### Behavioural Targeting
+Beyond demographic data, the recruitment strategy will aim to identify drivers who:
 
-**Pedestrians:**  
-- Parents, school staff, and local residents.
+- Drive frequently during peak hours near school zones.
 
-**Community Advocates:**  
-- Stakeholders involved in traffic safety initiatives.
+- Own or lease modified vehicles.
+
+- Exhibit risky behaviours, such as speeding or tailgating, based on self-reporting and survey data analysis.
+
+- Young males (18–34), a group overrepresented in speeding violations [@transportcanada2019].
 
@@
 **Multi-Stage Sampling:**
-- **Stratification by Wards:** Ensure geographic coverage across all Toronto wards with school zones.  
-- **Cluster Sampling:** Focus on areas near speed cameras and school safety zones.  
-- **Random Sampling:** Within clusters, randomly select participants for unbiased representation.
+
+The sampling approach ensures a balanced representation of speeding behavior across Toronto.
+Stratification by wards guarantees coverage of all areas with school zones, while cluster sampling targets locations near speed cameras and safety zones.
+Random sampling within clusters ensures fairness and unbiased participant selection.
 
@@
 Ads will use neutral language like "Toronto Traffic Safety Study" to avoid framing biases.
-  **Allocated Budget:** $200,000.  
-- **Community Partnerships:** Collaborations with schools, PTAs, and driving schools to involve families and new drivers.  
-  **Allocated Budget:** $50,000.
+Allocated Budget: $200,000.
+
+- **Community Partnerships:** Collaborations with schools, PTAs, and driving schools to involve families and new drivers. Allocated Budget: $50,000.
 
 ### Incentives
-- A raffle for ten $50 gift cards to encourage participation.  
-- Highlighting confidentiality ensures respondents feel safe sharing honest information.
-
-### Behavioral Targeting
-Beyond demographic data, the recruitment strategy will aim to identify drivers who:
-
-- Drive frequently during peak hours near school zones.
-
-- Own or lease modified vehicles.
-
-- Exhibit risky behaviors, such as speeding or tailgating, based on self-reporting and survey insights.
-
-- Young males (18–34), a group overrepresented in speeding violations [@transportcanada2019].
-
-- Drivers who own modified cars, as these factors are linked to increased risk of speeding [@safetyresearch2018].
+A raffle for ten $50 gift cards to encourage participation.
 
@@
 
-1. **Survey Naming:** The neutral title **Toronto Traffic Safety Study** reduces response bias by framing the study as a general exploration of driver behavior rather than focusing on speeding.
-2. **Confidentiality Assurance:** Respondents are explicitly informed that:
-   - Data is anonymous and used solely for research purposes.
-   - Identifiable information (e.g., email for raffle entries) is stored separately.
-3. **Question Design:** Questions are carefully ordered to build trust, with sensitive topics (e.g., fastest speed driven) placed later in the survey.
-4. **Response Validation:** Attention-check questions ensure participant engagement and accuracy.
+This study was strategically named to mitigate framing bias, presenting a neutral environment by focusing on general driver behavior rather than emphasizing speeding, which might have influenced responses.
+To build trust and encourage participation, confidentiality measures were implemented to avoid social desirability bias, with respondents assured that all data would remain anonymous and used solely for research purposes.
+Identifiable information, such as emails for raffle entries, was securely stored separately.
+The survey design also accounted for order bias, starting with general questions and gradually transitioning to more sensitive topics, such as the fastest speed driven.
+To address response bias and improve data accuracy, attention-check questions were included, ensuring participant engagement and validating the reliability of the responses.
+To minimize nonresponse and sampling bias, the study uses diverse recruitment channels, including targeted digital ads and community partnerships, alongside incentives like a gift card raffle to encourage broad participation.
+Stratification by wards ensures balanced geographic representation, avoiding overrepresentation of specific areas.
 
-## Data Validation
-1. **Logic Checks:**  
-   - Identify inconsistencies (e.g., reporting no awareness of cameras but frequent encounters).
+## Poll Aggregation
 
-2. **Cross-Validation:**  
-   - Compare survey responses with observational data (e.g., traffic camera footage and speed counts).
+A **poll-of-polls method** balances survey results with observational data.
+During analysis, survey responses will be weighted by proximity to enforcement measures and frequency of driving in school zones.
 
-3. **Geospatial Matching:**  
-   - Integrate survey responses with geospatial data from speed signs and cameras for location-specific analysis.
-
----
-
-## Poll Aggregation and Modeling
-### **Aggregation Approach**
-A **poll-of-polls method** balances survey insights with observational data:
-- Weight survey responses by:
-  - **Proximity to enforcement measures.**
-
-  - **Frequency of driving in school zones.**
-
-- Integrate responses with speed count data to build predictive models of speeding behavior.
-
----
-
-## Appendix Content
-1. **Survey Link:**  
+## Content
+**Survey Link:**  
    [https://forms.gle/TbRYwoGafy2EzwBQ9].
 
-2. **Full Survey Questions and Overview:**
+**Full Survey Questions and Overview:**
 
 **Title:** Toronto Traffic Safety Study  
-**Purpose:** To study driver behavior near school zones.  
+**Purpose:** To study driver behaviour near school zones.  
 **Estimated Time:** Less than 5 minutes.  
@@
 
-#### Section 1: Demographics and Background
+**Section 1: Demographics and Background**
 1. What is your age group?  
@@
 
-#### Section 2: Driving Habits
+**Section 2: Driving Habits**
 6. How often do you drive near schools in Toronto?  
@@
 
-#### Section 3: Speeding and Safety
+**Section 3: Speeding and Safety**
 9. What is the fastest you’ve ever driven (in km/h)? *(Select closest range)*  
@@
 
-#### Section 4: Traffic Safety Knowledge
+**Section 4: Traffic Safety Knowledge**
 13. **Test Question:** What is the legal speed limit near school zones in Toronto?  
@@
 
-#### Section 5: Traffic Incidents
+**Section 5: Traffic Incidents**
 14. Have you ever been involved in a collision while driving?  
@@
 
-#### Section 6: Raffle Entry
+**Section 6: Raffle Entry**
 16. Would you like to enter the raffle for a $50 gift card? *(Optional, email required)*  
@@
 
----
 # Residual Plot for model 1

### 2024-12-07 paper/paper.qmd
fix format, prose

@@
 ## Descriptive Statistics
-The descriptive statistics (see [@tbl-summary]) reveal important patterns within the dataset.
+The descriptive statistics in [@tbl-summary] reveal important patterns within the dataset.
 The average over-speeding value is 3.91 km/h, with extreme cases reaching as high as 70 km/h, illustrating the range of speeding behavior observed.
@@
 #| label: tbl-summary
-#| fig-cap: Summary of Joined Analysis Data
+#| tbl-cap: Summary of Joined Analysis Data
 #| echo: false
+#| warning: false
+#| message: false
 
@@
 summary_data %>%
-  kbl(caption = "Summary Statistics for the Final Filtered Data") %>%
+  kbl() %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
@@
     total_over_speeding = sum(over_speed_limit > 0, na.rm = TRUE),
-    avg_over_speeding = mean(over_speed_limit[over_speed_limit > 0], na.rm = TRUE)
+    avg_over_speeding = mean(over_speed_limit[over_speed_limit > 0], na.rm = TRUE), 
+    .groups = "drop"
   )
@@
 Other wards, such as Wards 13 and 16, also show elevated levels of over-speeding but fall significantly behind Ward 10.
-
 ```{r}
@@
 #| echo: false
-
+#| warning: false
+#| message: false
+#| 
 # Aggregate data for heatmap
@@
 Conversely, instances of extreme over-speeding (>50 km/h above the limit) are sparse but occur across wards 4, 13, and 25, highlighting potential hotspots for dangerous driving behaviour.
+
+## Interactive website
+An interactive Leaflet map was developed to visualize the locations of speed cameras and Watch Your Speed signs using a 500-meter spatial join.
+This map provides an accessible view of enforcement coverage across Toronto, allowing users to explore geographic distributions and the density of speed enforcement infrastructure.
+Markers represent the locations of speed cameras and signs showing the spatial reach of current traffic safety measures in school areas.
+Please see the website [here](https://karenrni.github.io/leaflet-web-map/).
 

### 2024-12-07 paper/paper.qmd
fix wording, add map info

@@
 ## Interactive website
-An interactive Leaflet map was developed to visualize the locations of speed cameras and Watch Your Speed signs using a 500-meter spatial join.
-This map provides an accessible view of enforcement coverage across Toronto, allowing users to explore geographic distributions and the density of speed enforcement infrastructure.
+An interactive Leaflet map was developed to visualize the locations of speeding incidents based on the speed cameras and Watch Your Speed signs using a 500-meter spatial join.
+This map provides an accessible view of enforcement coverage across Toronto, allowing users to explore geographic distributions and the density of speed enforcement infrastructure, as well as the speed limit, observed Speed (numeric midpoint of the speed bin), volume of incidents and ward number.
 Markers represent the locations of speed cameras and signs showing the spatial reach of current traffic safety measures in school areas.
